Adaptive evolution to thermal stress underpins climate resilience in a cosmopolitan arthropod

Version 1

Abstract

Adaptive evolution enables species to survive and thrive under changing environmental conditions. In the face of accelerating global climate change, thermal stress represents a major challenge to the persistence of terrestrial ectotherms. Understanding the genetic mechanisms underlying thermal adaptation is therefore critical for predicting species’ evolutionary potential and future success. Here, we combine experimental evolution, phenotypic assays, and multi-omics analyses to investigate the adaptive responses of the diamondback moth (Plutella xylostella), a globally destructive pest of cruciferous crops, to contrasting thermal environments. Populations evolved under hot (32°C/27°C) and cold (15°C/10°C) regimes exhibited distinct demographic and fitness traits relative to those maintained under favorable conditions (26°C). Integrated transcriptomic and metabolomic analyses revealed extensive transcriptional reprogramming and metabolic adjustments associated with temperature adaptation. Key pathways involved in lipid metabolism and DNA methylation were enriched, while functional validation using CRISPR-Cas9 confirmed that a mutant allele of PxSODC enhances superoxide dismutase activity and improves tolerance to thermal extremes. Together, these findings provides critical insights into this arthropod’s capacity for global dispersal and likely persistence under climate change, establishing a framework for understanding equivalent effects in other species.

Introduction

Human-induced climate change, particularly the continued change in temperature and precipitation patterns (IPCC, 2023), is affecting the survival and distribution of species (Fitzpatrick and Edelsparre, 2018), and impacts biodiversity and ecosystem services (Lawlor et al., 2024). Terrestrial insects are particularly vulnerable to temperature change because they are poikilothermic animals and small with high surface area to volume ratio (Wang et al., 2022). Over the past century, many insect populations have shown declining abundance, with climate change considered a major contributing factor after habitat loss and use of agricultural pesticides (Cardoso et al., 2020). It is predicted that a species distributed across latitudinal or elevational gradients shows a positive relationship between the thermal tolerance range and the level of climatic variability (Harvey et al., 2020). Tropical insects have a narrower thermal range, indicating that even a small increase in temperature can lead to a collapse in their populations (Harvey et al., 2022). Global climate change may also alter the geographical distribution of insect pests, allowing those previously confined by temperature barriers to spread to new areas, posing a threat to natural and managed ecosystems, crop production and food security (Deutsch et al., 2008; Outhwaite et al., 2022). Species with higher genetic diversity and adaptive potential may better cope with varying or extreme temperature environments, giving them an advantage to move to new habitats.

Adaptive evolution is a crucial mechanism for insects to tackle climate change and maintain or expand their geographical ranges (McCulloch and Waters, 2022; Burc et al., 2025). Understanding the genetic basis of thermal adaptation is essential for predicting how species will respond to rapid global climate change (Gibert et al., 2019). Genetic mutations, particularly non-synonymous mutations, are the primary sources of DNA sequence variability and can alter protein structure or function (Belfield et al., 2018). Increased thermal tolerance in organisms can originate from such mutations. For example, nonsynonymous mutations in the AcVIAAT gene of the eastern honeybee, Apis cerana (e.g., P42L substitution), are associated with enhanced thermal adaptation (Li et al., 2024c). During thermal evolution, alpine ground beetle, Nebria vandykei, can achieve survival capability in extreme thermal environments through adaptive selection of mutations in key genes (TREH, EIF3A, LRPPRC, etc.) coupled with immediate responses from heat shock proteins (Schoville et al., 2024). Although significant progress has been made in identifying genetic mutations that facilitate adaptation to extreme environments in non-model organisms, our understanding remains incomplete, especially regarding the integrated molecular, metabolic, and life history changes that collectively enable thermal adaptation.

In the case of diamondback moth (DBM), Plutella xylostella (Lepidoptera: Plutellidae), a globally serious pest of cruciferous plants, the availability of complete genome sequences and a large dataset of genome-wide single nucleotide polymorphism (SNP) sequences (You et al., 2013; You et al., 2020), allow for the investigation of its genetic plasticity, evolutionary history, and climate adaptation to changes in temperatures. Our recent research has shown that P. xylostella originated in South America and has expanded to all countries and regions where cruciferous crops are grown (Furlong et al., 2013; You et al., 2020). Based on the genomic data of field populations from 114 locations in a diverse range of biogeographical zones, we have demonstrated that P. xylostella is genetically capable of tolerating anticipated future climates in most regions of the world and will remain a global pest beyond 2050 (Chen et al., 2021). Here, we investigate the mechanisms of P. xylostella’s genetic adaptation and evolutionary responses to different thermal environments. Specifically, we employ thermal regime patterns of 12h/12h in hot (32°C/27°C) or cold (15°C/10°C) environments over the course of three years, compared to the favorable constant condition at 26°C, to investigate the adaptive evolution of P. xylostella in climate-controlled chambers.

Age-stage, two sex life tables of P. xylostella measured the demographic variation of the three P. xylostella strains evolved in the favorable (ancestral), hot and cold environments. Using metabolomic and transcriptomic analyses, we identified the key genes that could facilitate the adaptation of P. xylostella to thermal extremes. Our results showed that a large number of differentially expressed genes and metabolites were produced in populations adapted to high and low temperatures through multi-generational selection. We find the mutant of a key gene, PxSODC, which can alter the superoxide dismutase activity and increase the ability to scavenge superoxide anions, thereby significantly affecting the adaptability under high and low temperature environments. CRISPR-Cas9 was used to functionally validate the role of PxSODC in facilitating adaptive evolution and influencing regulatory networks. These findings not only improve our understanding of how novel thermal environments can drive genetically-based adaptive evolution manifested through integrated transcriptomic, metabolic, and life history changes in P. xylostella but also help to predict the distribution and dynamics of its populations in the context of global climate change.

Results

Life history trait divergence among temperature-adapted strains

Compared to the ancestral strain, both hot and cold strains had a significantly shorter preadult duration than the ancestral strain. The hot strain also have significantly shorter female longevity, and oviposition days, while the cold strain had an longer male longevity when compared to the ancestral strain (Supplemental Table S1). The female fecundity, intrinsic rate of increase (r) and finite rate of increase (λ) were all significantly higher in the hot strain than the cold strain which was not significantly different than the ancestral strain (Supplemental Table S1). Overall, following three years of evolution under contrasting thermal regimes, the hot and cold strains exhibited divergent life history traits compared to the ancestral strain. The hot strain exhibited accelerated life cycle and increased fecundity while the cold strain extended male longevity.

Based on the age-stage, two-sex life tables of the strains, we generated curves of the age-stage specific survival rate (sxj), age-specific survival rate (lx), female age-specific fecundity (fx), and population age-specific fecundity (mx) (Supplemental Figure S1). These curves present the dynamic variation in survival and fecundity of different strains and allow accurate estimation of daily survival and fecundity. The results showed that the daily survival rates at the pre-adult stages of the hot and ancestral strains were higher than that of the cold strain, with the maximum daily survival rate of HS = 78.9%, AS = 76.7% and CS = 66.7%. The daily fecundity of the hot and ancestral strains was higher than that of the cold strain, with the peak daily egg counts of HS = 53.22, AS = 53.00 and CS = 45.10 (Supplemental Table S1).

To assess the evolved thermal tolerance of the temperature-adapted strains, we further examined the stage-specific survival rates of the hot and ancestral strains under extremely high temperatures, as well as the supercooling and freezing points of the cold and ancestral strains at pupae stage. The survival rates of eggs, 3rd-instar larvae and adults in the hot strain were higher than those of the ancestral strain at 42°C, and the survival rate of pupae in the hot strain was higher than that in the ancestral strain at 43°C and 44°C (Figure 1A). The supercooling and freezing points of pupae in the cold strain were lower than those in the ancestral strain (Figure 1B). The variation in survival rates and the supercooling/freezing points at extreme temperatures suggest that the hot and cold strains of P. xylostella have undergone profound adaptive adjustments.

Evolved phenotypic changes in temperature-adapted strains.

(A) Stage-specific thermal tolerance responses (survival rates) of the ancestral and hot strains, with 20 individuals used in each of the four replicates for every treatment. (B) Supercooling and freezing points of pupae for the ancestral and cold strains, with 40 biologically independent samples used in each treatment. Data are presented as mean ± SEM. Statistical analyses are performed using t-tests with significant levels indicated by asterisks (*p < 0.05, **p < 0.01, ***p < 0.001).

Omics-based evidence for adaptive evolution

Our previous studies have identified metabolites such as trehalose and very long chain fatty acids that play a role in adaptation of P. xylostella to both high and low temperatures (Zhou et al., 2022; Lei et al., 2023). Here, we performed a broad analysis of targeted metabolites of 3rd-instar larvae of each strain using high-throughput UPLC-MS/MS. A total of 781 metabolites were identified, including 199 amino acids and their metabolites, 146 lipids, 90 organic acids and their derivatives, 78 nucleotides and their metabolites, 61 heterocyclic compounds, 45 benzene and substituted derivatives, 42 alcohols and amines, 37 carboxylic acids and derivatives, 21 coenzymes and vitamins, and 62 other metabolites (Figure 2A). Principal component analysis (PCA) and inter-sample correlation heat maps revealed significant metabolic changes in the 3rd-instar larvae from the ancestral strain to the hot and cold strains (Figure 2B; Supplemental Figure S2A). These comprised 77 differential metabolites (34 up-regulated, 43 down-regulated compared to the ancestral strain) in the hot strain, and 37 differential metabolites (13 up-regulated, 24 down-regulated compared to the ancestral strain) in the cold strain (Figure 2C). These differential metabolic profiles are likely to play crucial roles in the adaptation of P. xylostella to high and low temperature environments.

Metabolomic analysis of 3rd-instar larvae across ancestral, hot and cold strains (AS, HS and CS).

(A) Classification of metabolites, with a total of 781 metabolites being identified in different strains. (B) Principal component analysis (PCA) of the 781 metabolites across different strains. PC1 and PC2 represent the first and second principal components, respectively. (C) Volcano plot showing the down-regulated (green dots) and up-regulated (red dots) metabolites based on comparison between HS/CS and AS. (D) Classification of differential metabolites between HS/CS and AS. (E) Venn diagram showing the common and unique differential metabolites in HS and CS as compared to AS. (F) Fold changes and classifications of the common differential metabolites in HS and CS as compared to AS.

Compared to the ancestral strain, the common differential metabolites of the hot and cold strains included lipids, amino acids and their metabolites, organic acids and their derivatives, nucleotides and their metabolites, and benzene and substituted derivatives (Figure 2D; Supplemental Figure S2B). Inter-replicate analysis of differential metabolites showed a low correlation between the ancestral strain and hot/cold strains, but a high correlation between the hot and cold strains. Notably, 30 common metabolites were identified across the differential sets based on comparison of the hot/cold strains to the ancestral strain (Figure 2E; Supplemental Figure S2C). These metabolites, except for N1, N8-diacetylpiperidine, exhibited similar fold changes when comparing the hot/cold strains to the ancestral strain (Figure 2F). These results indicate that P. xylostella responds to different environmental stresses by regulating similar metabolic pathways. Further analysis revealed a reduction in most of the lipid metabolites by comparing the hot/cold strains to the ancestral strain (Figure 2F), providing evidence for adaptive evolution of P. xylostella in response to high and low temperatures, with the synthesis, degradation and metabolism of lipid playing an important role.

We then profiled and compared the transcriptomes of the three strains to identify the key genes involved in adaptation of P. xylostella to temperature extremes. This revealed significant variation in gene expression among strains (Figure 3A; Supplemental Figure S3A), with 1364 (825 up-regulated, 539 down-regulated) and 2029 (1205 up-regulated, 824 down-regulated) genes differentially expressed in the hot and cold strains, compared to the ancestral strain, respectively (Figure 3B). Pearson correlation showed, in contrast to the metabolomics data, a lack of strong correlation between the differentially expressed genes of the hot/cold strains and the ancestral strain (Figure 3C), with 498 common differentially expressed genes (Figure 3D). However, KEGG analysis revealed that the differentially expressed genes between the hot/cold strains and the ancestral strain were enriched in similar pathways, such as transport and catabolism, and the digestive system (Figure 3E), indicating that while multiple genes are involved in the adaptation of P. xylostella to high and low temperatures, a relatively limited range of biological functions might be affected.

Transcriptomic analysis of the 3rd-instar larvae across the ancestral, hot and cold strains.

(A) Principal component analysis (PCA) of genes across different strains. PC1 and PC2 represent the first and second principal components, respectively. (B) Volcano plots of differential gene expression, showing significantly up-regulated (red dots) and down-regulated (green dots) genes between HS/CS and AS (FDR < 0.05, fold change > 2). (C) Cluster analysis of the transcriptome, showing patterns of gene expression between different strains. (D) The number of common or unique differentially expressed genes between HS/CS and AS. (E) KEGG function classification of differentially expressed genes between HS/CS and AS.

The gene expression-based clustering tree using a weighted gene co-expression network analysis (WGCNA) was divided into 29 modules as shown with arbitrary color codes (Figure 4A; Supplemental Figure S3B). The turquoise module contained the most genes (3463), while the grey module contained the fewest (31) (Figure 4B). Selecting the common differential metabolites (30 in total) shared between the hot and cold strains as compared to the ancestral strain, we performed a correlation analysis with co-expressed networks and discovered the strongest correlations between genes in the red module and 28 metabolites (Figure 4C; Supplemental Figure S3C). Further analysis revealed that 79 genes within the red module were differentially expressed in the hot and cold strains when compared with the ancestral strain (Figure 4D). These results suggest that genes in the red module are most likely to play a critical regulatory role in the adaptation of P. xylostella to extreme temperatures.

Weighted gene co-expression network analysis (WGCNA) of transcriptomes for the 3rd-instar larvae across the ancestral, hot and cold strains (HS, CS and AS) of P. xylostella.

(A) Hierarchical cluster tree illustrating 29 modules identified by WGCNA. (B) Numerical distribution of genes of different modules as identified by WGCNA clustering. (C) In WGCNA, the red module shows the highest number of metabolites strongly correlated with genes. (D) Overlap of genes in the red module from WGCNA with common differentially expressed genes between HS/CS and AS.

Genetic basis of temperature adaptation

To further elucidate the genetic basis of P. xylostella adaptation to hot and cold environments, we analyzed 15 annotated genes with high expression levels (FPKM > 10) within the red module (Figure 4D) and identified 11 genes being successfully amplified. Comparative results revealed eight genes with nonsynonymous mutations and one with a synonymous mutation in both the hot and cold strains (Supplemental Table S2). Among these genes, we focused on the role of Px04C00666 (PxSODC) in the temperature adaptation of P. xylostella because the deletion of SOD genes can alter the response of insects to abiotic stresses including temperature (Bittner et al., 2019; Quan et al., 2024).

The NCBI database predicted that PxSODC contained three exons and two introns, with a conserved domain belonging to the copper-zinc superoxide dismutase superfamily (Supplemental Figure S4A) which plays an antioxidative role in cellular defense systems, protecting cells from damage caused by reactive oxygen species (Fridovich, 1995). The open reading frame of PxSODC was 633 bp and encodes 210 amino acids. Expasy predicted that the molecular weight of the PxSODC protein was 22,168.24 Da, with the isoelectric point being 6.29. Its secondary structure mainly consisted of strands, helices and coils (Supplemental Figure S4B). Evolutionary analysis showed PxSODC of P. xylostella clustered with that of other Lepidoptera insects such as Brenthis ino, Vanessa tameamea, and Vanessa cardui, indicating its conserved evolution within this taxonomic order, but distinct from Diptera flies and Hemiptera aphids (Supplemental Figure S4C). The coding region of PxSODC in the hot and cold strains had 23 SNP sites, including 20 synonymous and three non-synonymous mutations (Leu9-Val9, Lys25-Gln25, Leu194-Met194) (Figure 5A, 5B). Leu9-Val9 and Leu194-Met194 mutations were involved in the substitution of hydrophobic amino acids, with the Leu194-Met194 mutation exhibiting a higher frequency in both the hot and cold strains. The expression of PxSODC at different developmental stages of the hot and cold strains was significantly lower than that of the ancestral strain (Supplemental Figure S5A). After 2 h exposure of the 3rd-instar larvae to the stress of high (32°, 34°, 36°, 38° and 40°C) or low (12°, 10°, 8°, 6° and 4°C) temperature environments, the expression of PxSODC in the hot and cold strains was significantly lower than in the ancestral strain (Supplemental Figure S5B, S5C). This suggests that the PxSODC gene can alter its expression pattern and function in response to environmental change and contribute to the temperature-adaptive evolution in P. xylostella.

Role of PxSODC in temperature adaptation of P. xylostella.

(A) Allele frequencies of SNPs in the PxSODC gene amplified by PCR from the ancestral, hot and cold strains (AS, HS and CS). The analysis involves ten 4th-instar larvae from each of the strains; the dot (·) indicates identity with the reference base. (B) Frequency of amino acid translations from non-synonymous codon mutations in the PxSODC gene in different strains. (C) Stage-specific survival rates of the ancestral and mutant strains (AS, SODC-MU1, SODC-MU2 and SODC-MU3) under extreme heat conditions. (D) Supercooling and freezing points of the pupae from different strains (AS, SODC-MU1, SODC-MU2 and SODC-MU3). Data are presented as mean±SEM, one-way ANOVA with Tukey’s test was used for comparison.Six biologically independent samples were used in (C) and significant levels between groups with the same stress duration are indicated by asterisks (*p < 0.05, **p < 0.01, ***p < 0.001). A total of 40 biologically independent samples were used in (D) and statistical significance is indicated by different letters (p < 0.05).

To explore the specific function of PxSODC in temperature adaptation of P. xylostella, we successfully knocked out PxSODC in the ancestral strain using CRISPR/Cas9 technology. Of 162 eggs treated with CRISPR/Cas9, 75 successfully developed into adults. We confirmed three mutant strains in the G0 generation of P. xylostella: +1 bp (SODC-MU1), +2 bp (SODC-MU2) and −1 bp (SODC-MU3). Self-crossing continued, and three types of homozygous mutation were obtained in the G5 generation (Supplemental Figure S6). Age-stage, two sex life tables were constructed for the ancestral strain, and the three SODC-MU mutant (MU1, MU2, MU3) strains under constant favorable (26°C), hot (32°C/27°C) and cold (15°C/10°C) environments to examine their phenotypic traits including development, survival, fecundity and fitness (Supplemental Figure S7; Supplemental Tables S3-S5). Overall, the three mutant strains had a prolonged development time (T), lower survival rates (sxj and lx), and reduction in the fecundity (fx and mx) and population fitness parameters (r, λ, R0) compared to the ancestral strain. The observed change in these parameters indicated that the loss of PxSODC gene function significantly affected the life history and population fitness of P. xylostella, particularly in the hot/cold environments. Prolonged developmental time and reduced fecundity suggest that the physiological functions of the mutant strains may be damaged when exposed to both high and low temperature conditions.

We then carried out a detailed study on the response of ancestral and mutant strains (AS, SODC-MU1, SODC-MU2 and SODC-MU3) to extremely high temperature conditions. At 42°C, the survival rates of eggs, 3rd-instar larvae, female adults and male adults of the mutant strains were significantly lower from those of the ancestral strain at several time points (Figure 5C). However, survival rates of the mutant pupae exposed to the high temperature (43°C) were not significantly different from those of the ancestral strain at different time points (Figure 5C). The pupal stage exhibited enhanced stability to environmental temperature fluctuations compared to other life stages, possibly due to a special heat resistance mechanism in the pupae, such as protective chrysalis and U-shaped metabolism (Kaiser et al., 2010; Barros-Cordeiro et al., 2014). In addition, supercooling and freezing points of the mutant strains (MU1: −21.32±0.41 and −9.75±0.38; MU2: −21.50±0.38 and −9.93±0.43; MU3: −21.23±0.48 and −9.94±0.41) were significantly higher than those of the ancestral strain (−23.09±0.26 and −11.58±0.52) at pupal stage (Figure 5D), indicating a key role of the PxSODC gene in the adaptability and tolerance of P. xylostella to extreme temperatures.

To further investigate the effect of PxSODC gene mutations on the temperature adaptability of P. xylostella, we identified five genes from the same SOD family in transcriptomes of the 3rd-instar larvae from the three tested strains. We found that Px04C00505 and Px13C00423 showed SNP mutations in the hot and cold strains, whereas Px20C00248, Px15C00224 and Px15C00223 were not mutated (Supplemental Table S6). Further comparison of gene expression across different strains revealed that, relative to the ancestral strain, the expression levels of PxSODC, Px04C00505, and Px13C00423 were significantly reduced in the hot and cold strains, while the remaining genes maintained stable or increased expression levels (Figure 6A). Concurrently, SOD activity decreased in the hot and cold strains, along with a reduction in O2- levels (Figure 6B, 6C). When SOD expression and activity, as well as O2- levels, were compared under different temperature conditions between the ancestral and hot/cold strains, similar patterns were observed (Figure 6D, 6E, 6F). These results demonstrate that the hot and cold strains could increase their O2-scavenging capacity through non-synonymous mutations, thereby better adapting to high or low temperature environments. Compared to the ancestral strain, expression of the mutated SODC genes (Px04C00505 and Px13C00423) was increased in the male adults in SODC-MU1 and SODC-MU2. The expression levels of the non-mutated genes Px20C00248, Px15C00224 and Px15C00223 also increased (Figure 6G). Further, SOD activity decreased, while O2- levels increased in the two mutant strains (Figure 6H, 6I), which were unable to fully compensate for the effects caused by the deletion of the PxSODC gene, implying that the SOD protein encoded by PxSODC plays a crucial role in O2- scavenging.

SOD expression and activity and superoxide anion (O2-) levels across developmental stages and temperature environments in different strains of P.

xylostella. (A) Expression levels of the SOD genes at different developmental stages of AS, HS, and CS in the favorable temperature environment (26°C). (B) SOD enzyme activity at different developmental stages of the ancestral, hot and cold strains (AS, HS, and CS) in the favorable temperature environment. (C) O2- levels at different developmental stages of AS, HS, and CS in the favorable temperature environment. (D) Expression levels of the genes from the SOD family in the 3rd-instar larvae of AS, HS, and CS in the hot (32°C, 34°C, 36°C) and cold (12°C, 10°C, 8°C) environments. (E) SOD enzyme activity in the 3rd-instar larvae of AS, HS, and CS in the extreme temperature environments. (F) O2- levels in the 3rd-instar larvae of AS, HS, and CS in the hot and cold environments. (G) Expression levels of SOD family (excluding the PxSODC gene) at different developmental stages of the ancestral and SODC-MU strains in the favorable temperature environment. (H) SOD enzyme activity at different developmental stages of the ancestral and SODC-MU strains in the favorable temperature environment. (I) O2- levels at different developmental stages of the ancestral and SODC-MU strains in the favorable temperature environment. n = 3 biologically independent samples in (A), (D), (G); within each of the boxes, the numerical value represents log2-fold change of the gene expression level in the treated samples with respect to the control. n = 4 biologically independent samples in (B), (C), (E), (F), (H), with data being presented as mean±SEM. One-way ANOVA with Tukey’s test was used for comparison in (A), (B), (C) and (G), (H), (I) (p < 0.05). t-test was used for comparison in (D), (E), (F) (p < 0.05).

PxSODC-allied metabolic networks

To explore the metabolic networks associated with the PxSODC gene in P. xylostella and better understand the biological functions underlying their complex relationships, we performed an untargeted metabolomic analysis on the ancestral and mutant (SODC-MU1 and SODC-MU2) strains at different developmental stages. We detected 608 metabolites across the strains, including 167 organic acids and their derivatives, 102 nucleotides and their metabolites, 102 lipids, 92 amino acids and their metabolites, 33 heterocyclic compounds, 24 carbohydrates and their metabolites, and 88 other metabolites (Supplemental Figure S8A). Principal component analysis (PCA) revealed a distinction between the ancestral and mutant strains along the first component (PC1). However, a trend of clustering was observed in the 3rd-instar larvae of ancestral and SODC-MU1 strains and the male adults of ancestral and SODC-MU2 strains, suggesting that the difference at these developmental stages may be driven by a few key metabolites (Supplemental Figure S8B). Compared to ancestral strain, different metabolites were identified in the eggs, 3rd-instar larvae, pupae, female adults and male adults of mutant strains (Supplemental Figure S8C). The mutant strains had 68, 103, 110, 77 and 35 common differential metabolites compared to the ancestral strain at different developmental stages of P. xylostella (Supplemental Figure S8D). In addition, the lipid content significantly decreased in the 3rd-instar larvae, pupae and female adults, while the content of nucleotides and their metabolites increased in the eggs, 3rd-instar larvae and pupae following deletion of the PxSODC gene in P. xylostella (Supplemental Figure S9A). The up-regulation and down-regulation of various metabolites may be associated with the antioxidant function of PxSODC. We observed that the deletion of PxSODC gene could lead to elevated levels of the O2- within P. xylostella (Figure 6I).

We then performed enrichment analysis of differential metabolite KEGG pathways for the SODC-MU (MU1 and MU2) eggs, 3rd-instar larvae, pupae, female adults and male adults resulting in these differential metabolites wre distributed in 23, 28, 28, 22, and 18 pathways, respectively. They were mainly involving lipid metabolism, nucleotide metabolism, carbohydrate metabolism, cofactor and vitamin metabolism, and amino acid metabolism (Supplemental Figure S9B). These pathways were related to the PxSODC gene and might have contributed to the temperature-adaptive capacity in P. xylostella through the regulation of biological functions/processes in different temperature environments.

In the metabolome, we observed that the abundance of 5-hydroxymethyluracil, 2-methylcitric acid and 5’-deoxyadenosine differed in all developmental stages of mutant strains (SODC-MU1 and SODC-MU2) compared to the ancestral strain (Figure 7A). Apart from a significant increase in 5’-deoxyadenosine in the 3rd-instar larvae of mutant strains compared to the ancestral strain, the levels of these three metabolites were significantly reduced in other developmental stages, suggesting a possible direct association with PxSODC (Figure 7B) and representing a key biological regulatory response adopted by P. xylostella in adaptation to different environmental conditions. The significant roles of 5-hydroxymethyluracil (involved in dynamic methylation regulation) and 5’-deoxyadenosine (as a precursor to methyl donors) in DNA methylation has been previously confirmed (Bhutani et al., 2011; McKean et al., 2019). We measured the expression levels and enzymatic activities of DNA methyltransferase 1 (PxDnmt1) in the 3rd-instar larvae of different strains. The results showed that both the expression and enzymatic activity of PxDnmt1 were significantly reduced in the hot and cold strains compared to the ancestral strain (Figure 7C, 7D). Using RNA interference (RNAi) technology, we specifically silenced the expression of PxDnmt1 in the ancestral strain of P. xylostella (Figure 7E) and observed significantly reduced levels of 5-methylcytosine (5-mC, a marker of DNA methylation) in both pupae and female adults (Figure 7F) (Ni et al., 2023). Further, we found that silencing of PxDnmt1 could significantly decrease the critical thermal maximum (CTmax) of female adults and increase the supercooling and freezing points of pupae (Figure 7G, 7H). These results confirm that DNA methyltransferase is associated with the adaptive response of P. xylostella to extreme temperatures.

Comparison of metabolites and DNA methylation across different strains of P. xylostella.

(A) A Venn diagram showing the metabolites that are consistently different between the ancestral and mutant strains across different developmental stages. (B) Three metabolites with persistent discrepancy across different developmental stages in the ancestral and mutant and strains. (C) Expression level of the DNA methyltransferase 1 gene (PxDnmt1) in the ancestral, hot and cold strains. n = 16 biologically independent samples. (D) DNA methyltransferase activity in the ancestral, hot and cold strains. n = 12 biologically independent samples. Data are presented as mean±SEM, one-way ANOVA with Tukey’s test was used for comparison in (C), (D) (p < 0.05). (E) Injection of dsDnmt1 significantly reduced the expression level of PxDnmt1 in the ancestral strain of P. xylostella. n = 3 biologically independent samples. (F) Silencing of PxDnmt1 decreased 5-methylcytosine (5-mC) content in the female adults and pupae of P. xylostella. n = 4 biologically independent samples. (G) Female adults with silenced PxDnmt1 exhibited a significantly decreased critical thermal maximum (CTMax). n = 20 biologically independent samples. (H) Pupae with silenced PxDnmt1 displayed elevated supercooling and freezing points. n = 40 biologically independent samples. Data are presented as mean±SEM, unpaired t-test was used for comparison in (E), (F), (G), (H) (p < 0.05).

Discussion

Insects are valuable bio-indicators of the effects of climate change via their phenology, distribution and species interactions, population dynamics and community structure responses (Bale et al., 2002; Halsch et al., 2021). The present study demonstrates that P. xylostella can undergo rapid genetic adaptation to thermal extremes, with evolved changes spanning transcriptomic, metabolomic, and life history traits. By conducting the life history trait, demography, and fitness assessment of the ancestral, hot and cold strains of P. xylostella, we observed that the hot and cold strains had evolved significant genetic differences from the ancestral strain in multiple traits related to thermal tolerance and fitness. In the context of global warming, P. xylostella may evolve greater flexibility across its range and ecological niche leading the hot-evolved populations to be able to persist in regions with increased temperatures due to climate change (Chen et al., 2021). With climate change, cold-adapted P. xylostella may be favored during episodes of late frost in the spring or early frost in the autumn in temperate regions. The lower supercooling and freezing points of the cold strain pupae facilitate the survival of insects in cold climates, extending their ecological adaptability to low-temperature environments (Block, 1997). The demonstrated capacity of P. xylostella to adapt to extreme thermal conditions of both forms implies that P. xylostella may survive under a broader range of climatic conditions, posing new challenges for the management and control of this worldwide pest.

Our findings reveal a significant metabolic adjustment in P. xylostella during its adaptive evolution to both high and low temperatures. Lipids such as octadecenoic acid, epoxystearic acid and carnitine are reduced in the hot and cold strains compared to the ancestral strain, which indicates that after exposing and adapting to novel temperature environments, P. xylostella may maintain basic metabolic rates by lowering the level of certain metabolites, thereby allocating more energy to development and reproduction (Mallard et al., 2018). Lipids play a key role in energy storage and membrane stability in insects; adjusting the lipid metabolism may be a common mechanism for insects to adapt to temperature changes (Rommelaere et al., 2019). Carnitine has important functions in fatty acid transport and energy metabolism; change in its levels may impact energy supply (Bremer, 1983). Although gene expression between the hot and cold strains does not show a strong correlation with metabolites, KEGG analysis indicates that differentially expressed genes between the ancestral strain and the hot or cold strains have similarly enriched pathways. More differentially expressed genes, but fewer differential metabolites, were found in the cold strain compared to the ancestral strain than in the hot strain, a phenomenon different from that observed in Drosophila (Mallard et al., 2018). This might be due to the species-specific difference or the difference in selection pressure in the cold environment and requires further investigation.

Integrating information from metabolomic and transcriptomic analyses revealed extensive changes in gene expression and metabolic rewiring, suggesting that P. xylostella could adopt similar adaptive evolutionary strategies in hot and cold environments. Although the specific genes involved in the adaptation may be different, their functional pathways are similar, especially the important role of lipid metabolism in temperature adaptation, possibly affecting the environmentally sensitive phenotypes of heat tolerance, cold hardiness, and female fecundity (Tigano et al., 2020; Sherpa et al., 2022; Li et al., 2024a).

Insect genomes contain genes encoding the proteins involved in thermal adaptation (Kelley et al., 2014). Mutations in these genes during adaptive evolution in response of thermal adaptation can lead to novel phenotypic traits, such as changes in the life history and population fitness of insects (Gibson et al., 2019). The PxSODC gene encodes a superoxide dismutase that scavenges superoxide anions in cells, maintaining redox balance and protecting cellular structures (Sheng et al., 2014). Under extreme temperatures, insects can adjust their survival strategy, allocating more energy to maintain fundamental life functions (Hoffmann and Sgro, 2011). Here, the deletion of the PxSODC gene led to reduced SOD enzyme activity and increased O2- levels, reducing the tolerance of P. xylostella to extreme temperatures. These findings provide novel insights into how genetic variation translates into phenotypic variation (You et al., 2024), and the ways in which P. xylostella responds – and has responded – to a changing climate (Chen et al., 2021).

Expression of the PxSODC gene in both hot- and cold-evolved strains was lower than that in the ancestral strain, and the two evolved strains also had lower O2- levels, suggesting that they should have adapted to extreme temperature environments. This adaptation may result from non-synonymous mutations that alter the conformation of the PxSODC-encoded protein, increasing its capacity to scavenge superoxide anions and reducing the demand for SOD enzymes. A previous study has also shown that mutations in the Mn-SODC protein at positions K182R and A183P can increase its enzyme activity (Sheng et al., 2013). After long-term adaptation, insects may acquire the ability to maintain cellular homeostasis in new thermal environments by reducing basal metabolism and allocating more energy to development and reproduction (Mallard et al., 2018). At 34°C and 36°C, the trends in SOD enzyme activity and PxSODC gene expression differ between different temperature-evolved strains, suggesting the involvement of additional genes in the regulation of SOD enzyme activity. This hypothesis is supported by our transcriptomic analysis identified additional SOD genes (Supplemental Table S5).

In the SODC gene family, two SOD genes (Px04C00505 and Px13C00423) underwent non-synonymous mutations and showed reduced expression in the hot and cold strains at different developmental stages, while the expression of three SOD genes without mutations remained relatively stable or increased at most developmental stages. The evolution of protein functions is driven by mutations, which in some cases can switch directly from one function to another through single amino acid changes (Nobeli et al., 2009; Patsch et al., 2024). Under extreme temperatures, the expression of these genes in the 3rd-instar P. xylostella larvae of the hot and cold strains trends to be similar to those of the ancestral strain. This indicates that adverse environmental conditions increase intracellular oxidative stress, which requires regulation of SOD expression and enzyme activity to scavenge superoxide anions (Islam et al., 2022). Maintaining high levels of SOD enzyme activity requires additional energy, placing a strain on cellular energy metabolism and resource allocation (Emre et al., 2013). SOD genes with non-synonymous mutations, similar to PxSODC, can lead to the change in protein structure or function, affecting enzyme activity and allowing for faster O2- scavenging at lower transcript levels, reducing resource requirements. The three unmutated SOD genes, if mutated, might adversely affect the moth or have other functions, such as involvement in cellular signaling pathways (Mondola et al., 2016).

Metabolomic analysis of different developmental stages in the ancestral and mutant strains revealed that after the loss of PxSODC gene, the metabolism of P. xylostella underwent temperature-adaptive adjustments involving lipids, nucleotides, carbohydrates, coenzymes and vitamins, and amino acids. This study also showed that DNA methylation plays a key role in the temperature adaptation of P. xylostella. While DNA methylation may be associated with gene activation, its main function remains the inhibition of gene expression (Stroud et al., 2015; Wang et al., 2018). Transcriptomic analysis also showed that more genes were up-regulated in the hot and cold strains compared to the ancestral strain, highlighting the role of DNA methylation in regulating gene expression to help P. xylostella maintain physiological functions and survive. In addition to directly regulating the expression of temperature responsive genes, epigenetic effects can also indirectly affect the response of insects to temperature challenges (Reynolds, 2017). By adding the molecular data of epigenetic markers, underlying mechanisms on the adaptive adaptation can be more easily elucidated. Therefore, further work is required to better understand the impact of non-genetic effects on adaptation to future climates including how they interact with genetic adaptive capacity.

This study elucidates the molecular mechanisms underlying adaptation of Plutella xylostella to both high- and low-temperature environments and functionally validates differentially expressed genes identified in ancestral, hot-evolved, and cold-evolved strains. Nevertheless, thermal adaptation in arthropods may engage distinct, temperature-specific biological pathways; accordingly, future work will prioritize the characterization of strain-specific differentially expressed genes. Beyond functional validation of the canonical stress-associated gene PxSODC, additional genes harboring nonsynonymous mutations warrant detailed investigation to clarify their roles within the broader regulatory network. Importantly, our findings also underscore a critical role for DNA methylation in thermal adaptation in P. xylostella. Building on this insight, we plan to generate PxDnmt1 knockout lines and apply whole-genome bisulfite sequencing (WGBS) across thermal strains to identify key methylation sites, thereby enabling a more comprehensive assessment of their contributions to temperature adaptation in P. xylostella and the generality of these mechanisms across other species.

Materials and methods

Insects

The founding population of P. xylostella was established from field-collected larvae on cabbage (Brassica oleracea var. capitata) in organic farms in Fuzhou, Fujian Province, China (26°05’N, 119°18’E) in July 2012. The collection site was confirmed through farmer interviews and local agricultural records to have no history of insecticide application for at least five years. Approximately 300-500 larvae were collected from multiple host plants distributed across a 2-hectare area to maximize genetic diversity and minimize founder effects. The field-collected population was reared in the laboratory under controlled conditions of 26°C and 60% relative humidity with a 12-h light/12-h dark cycle, without exposure to insecticides. This setup was referred to as the favorable temperature environment based on a previous study on the relationship between temperature and the P. xylostella development (Liu et al., 2002). Population size was maintained at >500 individuals per generation to minimize inbreeding and genetic drift. This laboratory-acclimated population was designated as the ancestral strain and served as the baseline for all subsequent experiments. Eggs and larvae were reared in sterile plastic disposable Petri dishes (90 mm) with an artificial diet containing 68 g yeast powder, 20.4 g agar, 127.5 g raw wheat germ, 3.4 g potassium sorbate, 3.4 g methyl paraben, 34 g sucrose, 10.2 g powder of radish seed, 1.7 g vitamin premix, 3.4 g ascorbic acid, 3.4 mL cola oil, 0.34 mL linoleic acid and 850 mL water The adults were allowed to mate and lay eggs in disposable paper cups, and were fed with a 10% honey solution.

The previous study on the relationship between temperature and developmental rate shows that P. xylostella can survive and develop at the temperatures 32°C at the maximum, 26°C as the optimum, and 10°C for the minimum (Liu et al., 2002). To generate thermally adapted populations, we established 18 replicate populations from the ancestral strain and randomly assigned them to three thermal regimes: (1) a hot-evolved treatment with cycling temperatures of 32°C/27°C (12 h light/12 h dark), (2) a cold-evolved treatment with cycling temperatures of 15°C/10°C (12 h light/12 h dark), and (3) a control treatment maintained at constant 26°C. All other environmental conditions (humidity, photoperiod, diet) remained identical across treatments. The populations were maintained with non-overlapping generations and a census population size of approximately 500 individuals per replicate.

The thermal selection experiment was conducted for approximately three years. To minimize maternal effects and ensure that observed differences were due to genetic adaptation rather than developmental plasticity, all populations were reared for two additional generations under common garden conditions (26°C) prior to phenotypic and molecular assays. For each treatment, we randomly selected six replicate populations for downstream experiments, resulting in 18 experimental cohorts in total.

Development of age-stage-specific two-sex life tables

To study the life history and population fitness of the three different P. xylostella strains, age-stage-specific life tables were developed. A total of 90 newly laid eggs from each strain were transferred to three 90 mm diameter Petri dishes (30 eggs per dish) (Chi et al., 2020).

The stage-specific number of individuals was recorded daily (food replaced every two days). During the pupal stage, each pupa was placed in a perforated 1.5 mL centrifuge tube. After eclosion, adults were transferred to 50 mL plastic cups for mating and oviposition and fed with a 10% honey solution. We continuously monitored and recorded daily oviposition and the number of surviving adults within the population until all individuals died.

For the individual life tables, 120 randomly selected newly laid eggs from each strain (i.e., 120 eggs/strain/temperature) were placed under different temperature conditions (favorable, hot, and cold environments). The eggs were individually transferred to 30 mm diameter petri dishes (artificial diet replaced every two days). The number of surviving larvae and pupae was recorded daily, and each pupa was placed individually in a perforated 1.5 ml centrifuge tube. One newly emerged adult female and one newly emerged adult male were placed in a 25 mL plastic cup and fed with a 10% honey solution for mating and oviposition. The number of eggs laid by each adult female was recorded daily until death. The longevity of both males and females was recorded.

Life history and population fitness parameters were calculated to assess the life history traits of P. xylostella evolved in different temperature environments based on the recorded data of age-stage-specific two-sex life tables of different strains. The numerical computation was done using the TWOSEX-MSChart software (http://140.120.197.173/ecology/prod2.htm) and the bootstrap method with 100,000 replications to obtain standard errors of the fitness parameters. The paired BT method was also used to assess the difference between the different treatments. A P value of less than 0.05 indicates a statistically significant difference (Chi, 1988).

Metabolomic and transcriptomic profiling

Samples collected for metabolomic profiling included: (1) the 3rd-instar larvae from different strains (AS, HS and CS) maintained at the favorable temperature (26°C); and (2) the 2-day-old eggs, 1-day-old 3rd-instar larvae, 2-day-old pupae and newly emerged adults from the ancestral and SODC-MU (MU1 and MU2) strains (see section Deletion of the targeted gene using the CRISPR/Cas9 genome editing) at the favorable temperature. Six biological replicates were used for metabolomic profiling and three replicates for transcriptomic profiling for each of the strains.

Metabolomic profiling

Stored samples were thawed on ice and weighed (50 ± 2 mg) into 1.5 mL centrifuge tubes, to which three pre-cooled steel balls (3 mm) and 500 µL of pre-cooled 70% methanol (Merck, Germany) were added. Each of the samples was homogenized in a pre-cooled tissue homogenizer (25 HZ, 5 min) (Tissuelyser, Qiagen), and then the homogenized sample was left on ice for 15 min, then centrifuged at 12,000 rpm at 4°C for 10 min. The supernatant was transferred to a new 1.5 mL centrifuge tube and stored overnight at −20°C. The following day, the samples were centrifuged at 12,000 rpm for 3 minutes at 4°C. The supernatant was then collected with a sterile syringe and filtered through a 0.22 µm filter (Waters, USA) into an HPLC sample vial. The instrumental system for data acquisitions mainly used ultra-high performance liquid chromatography and tandem mass spectrometry (multiple reaction monitoring mode). The chromatography and tandem mass spectrometry conditions were as described by Li et al. (2021) Significant metabolites were identified using the criteria: |log2Fold Change| ≥ 1; VIP ≥ 1; P < 0.05. All differential metabolites were assigned to different pathways by KEGG analysis.

Transcriptomic profiling

mRNA libraries were constructed for each of the samples and sequenced on the Illumina HiSeq4000 platform (Illumina, San Diego). Raw reads obtained from sequencing were filtered, low quality reads were removed using adapters, and clean reads were obtained for subsequent information analysis. Clean reads were aligned to the P. xylostella genome using HISAT2 (V2.1.0) (http://121.37.197.72:5010), with sequence alignment performed using the software’s default parameters. Gene expression levels were measured using FPKM (fragments per kilobase of transcript per million fragments mapped). The P-value was corrected for multiple hypothesis testing following Benjamini and Hochberg (1995). Differential expression analysis between samples was performed using DESeq2. Differentially expressed genes were identified using the criteria of |log2Fold Change| ≥ 1 and FDR (false discovery rate) < 0.05. All differentially expressed genes were classified into different pathways by KEGG analysis.

Weighted gene co-expression network analysis (WGCNA)

In the transcriptome data set, genes with an FPKM value below 0.1 were filtered out. The WGCNA package in R (v2.0) was used to calculate weight values and a soft threshold was determined based on the scale-free network principle. A gene clustering tree was constructed based on gene expression correlations, and gene modules were identified based on these clustering relationships. The minimum number of genes per module was set to 30, and the cut height threshold was set to 0.25 to merge potentially similar modules, with other parameters set to default values. Trait data, comprising differential metabolites between the hot and cold strains relative to the ancestral strain, were used for correlation analysis. Correlation was considered to be significant if the absolute correlation coefficient was greater than 0.8 and P < 0.05. Genes within specific modules were compared with differentially expressed genes and those common to both were considered as candidate genes.

RNA extraction and cDNA synthesis

Total RNA was extracted using the Eastep® Super Total RNA Extraction Kit (Promega, Shanghai) according to the manufacturer’s instructions. RNA integrity and quality were assessed using a NanoDrop 2000 spectrophotometer (GE Healthcare, USA) and 2% agarose gel electrophoresis. Total RNA (2000 ng) was reverse transcribed into cDNA using the Reverse-Transcription System Kit (Promega, Shanghai).

Gene cloning

The following gene sequences were amplified: (1) differentially expressed genes identified from specific modules; and (2) genes from the PxSODC gene family. Reference sequences for these genes were obtained from the P. xylostella genome database and full-length primers were designed using Primer Premier 6.0 (Supplemental Table S7). PCR amplifications were performed using cDNA from the ancestral, hot and cold strains of P. xylostella (at least six samples) as templates, using Phanta® Max Super-Fidelity DNA Polymerase (Vazyme, China).

The 50 µL PCR reaction mixture contained 2× reaction buffer (25.0 µL), dNTP mix (1.0 µL), 20 µM upstream primer F (2.0 µL), 20 µM downstream primer R (2.0 µL), DNA polymerase (1.0 µL), nuclease-free water (15.0 µL) and cDNA (4.0 µL). Amplification conditions were as follows 95°C for 3 minutes, followed by 35 cycles of 95°C for 15 seconds, 58°C for 15 seconds and 72°C for a gene-specific extension time, with a final extension at 72°C for 5 minutes. The integrity of the PCR products was verified by 2% agarose gel electrophoresis and purified using the Omega gel extraction kit (USA). The purified fragments were cloned into the pESI-Blunt vector using the Hieff CloneTM Zero TOPO-Blunt Simple Cloning Kit (TIANGEN, China) and transformed into competent DH5α cells (YEASEN, Shanghai). The ligated product was transferred to DH5α competent cells and plated on LB+100 μg/ml ampicillin plates and incubated overnight at 37°C. A single colony was picked and placed in 500 µL of liquid LB with 100 µg/mL ampicillin. A positive clone was sent to Sangon Biotech (Shanghai, China) for sequencing. Sequence alignments of candidate genes from different strains were performed using Snap Gene software.

Sequence analysis and phylogenetic tree construction

To determine the exons and introns of the target genes, their sequences were aligned with gDNA from the P. xylostella genome database. Protein sequences were analyzed using a protein sequence analysis and classification tool (InterPro, http://www.ebi.ac.uk/interpro/). The relative molecular masses and isoelectric points of the proteins were predicted using Expasy (http://expasy.org/tools/dna.html). The secondary structures of the proteins were predicted using PSIPRED (http://bioinf.cs.ucl.ac.uk/psipred). A phylogenetic tree was constructed using the neighbor-joining method in MEGA X software, and the robustness of the tree was verified by bootstrap analysis (bootstrap = 1000 replicates).

qRT-PCR analysis

For real-time quantitative PCR analysis, we collected samples as follows: (1) we randomly selected 22 genes from the transcriptomes of the 3rd-instar larvae of ancestral, hot, and cold strains to validate the transcriptome data; (2) we collected samples from the eggs, 3rd-instar larvae, pupae and adult males and females of the ancestral, hot, cold and SODC-MU (MU1 and MU2) strains to assess the transcription levels of genes including PxSODC, Px04C00505, Px13C00423, Px20C00248, Px15C00224 and Px15C00223; (3) we collected the 3rd-instar larvae of the ancestral and hot strains exposed to different high temperature treatments (32°C, 34°C, 36°C) to analyze the transcription levels of the above-mentioned genes; and (4) we collected the 3rd-instar larvae of the ancestral and cold strains exposed to different low temperature treatments (12°C, 10°C, 8°C) for similar assessments.

qRT-PCR primers were designed using Primer Premier 6.0, with PxRPL32as the reference gene (Supplemental Table S7). The reaction mixture of 20 μL contained 10 μL 2× Taq Pro Universal qPCR Master Mix (Vazyme, Nanjing, China), 0.4 μL of each primer, 7.15 μL nuclease-free water and 2.0 μL cDNA. The QuantStudio Real-Time PCR System (Thermo, USA) protocol was as follows: 95°C for 30 s; 40 cycles of 95°C for 10 s, 60°C for 30 s; followed by melting curve analysis of 95°C for 15 s, 60°C for 1 min, 95°C for 15 s. Each sample contained 3 biological replicates and 3 technical replicates, and the relative expression of genes was calculated using the 2-ΔΔCt method.

Deletion of the targeted gene using the CRISPR/Cas9 genome editing

We generated stable homozygous mutant strains of P. xylostella with the PxSODC gene deleted using the CRISPR/Cas9 system. The target site was designed based on the 5’-N20NGG-3’ motif (underscore indicates PAM sequence), and the potential off-target effect of sgRNAs was predicted using Cas-OFFinder (http://www.rgenome.net/cas-offinder). The in vitro transcription template for sgRNA was generated from a single nucleotide strand under the following conditions: 95°C for 3 min, followed by 35 cycles of 95°C for 15 s, 68°C for 15 s and 72°C for 30 s, with a final extension at 72°C for 5 min. The amplified product was purified by gel extraction. The sgRNA was obtained by in vitro transcription of the gel-purified product using the HiScribe™ T7 Quick High Yield RNA Synthesis Kit (New England Biolabs, USA). The reaction mixture contained 2.5 μL NTP buffer mix, 0.5 μL T7 RNA polymerase mix, 65 ng gel-purified product, made up to 5 μL with nuclease-free water. After overnight incubation at 37°C, 0.5 μL of DNase was added to remove DNA and the product was incubated at 37°C for 20 minutes to yield sgRNA. The sgRNA was purified by phenol-chloroform extraction and stored at −80°C.

We prepared a 10 μL reaction mixture containing 300 ng/μL sgRNA and 200 ng/μL Cas9 protein (GenCrispr, Nanjing), 1 μL 10× reaction buffer and nuclease-free water to make up to 10 μL and incubate at 37°C for 25 minutes. The mixture was injected into freshly laid eggs using the Olympus SZX16 microinjection system (Olympus, Japan) and the entire microinjection was completed within 30 minutes of eggs being laid. After injection, the eggs were placed in a Petri dish and the number of eggs hatched was recorded. Adult gDNA was extracted using the TiANamp Genomic DNA Kit (TIANGEN, China). Specific primers were designed for PCR amplification (Supplemental Table S7), with conditions as follows: 95°C for 3 min, followed by 34 cycles of 95°C for 15 s, 58°C for 15 s and 72°C for 15 s, with a final extension at 72°C for 5 min. The sequence of the PCR products was checked by Sangon Biotech (Shanghai) Co., Ltd.

The injected eggs were referred to as the G0 generation. These were reared to adulthood, crossed with the ancestral (non-injected) adults and used to extract genomic DNA from G0 adults after oviposition (the resulting progeny representing the G1 generation). PCR products flanking the two sgRNA target sites were amplified as mentioned above to determine genotypes and identify heterozygotes (individuals with double peaks in the sequence chromatogram starting from the sgRNA target site). The G1 generation was self-crossed to produce the G2 generation, and all G1 adults were genotyped based on PCR amplification for individual identification. G2 progeny derived from G1 heterozygotes with the same allelic mutation were selected. The G2 generation was then self-crossed to produce the G3 generation, retaining those with the same type of homozygous mutations to establish homozygous lines. If the G3 generation remained heterozygous, self-crossing continued until homozygous mutations were obtained (Wang et al., 2020). By the end, three mutants were obtained and called SODC-MU (MU1, MU2 and MU3) strains.

RNA interference mediated silencing of target genes

Gene-specific primers containing T7 promoter sequences were designed (Supplemental Table S7), and PCR was performed using total P. xylostella cDNA as a template. The PCR products were purified using a gel extraction kit. Double-stranded RNA (dsRNA) was synthesized by in vitro transcription using the TranscriptAid T7 High Yield Transcription Kit (Vazyme, Nanjing, China). The dsRNA was diluted to 2 μg/μL using DEPC-treated water (Beyotime, Shanghai, China). A volume of 500 nL diluted dsRNA (dsEGFP or dsDnmt1) was injected into pupae using a Nanoliter 2000 microinjector (World Precision Instruments LLC, USA). Total RNA was extracted 24 hours after injection and reverse transcribed to cDNA. Gene knockdown efficiency was analyzed by qPCR using beetles injected with dsEGFP as controls. The experiment was performed in three independent biological replicates (Zhou et al., 2024).

Assessing the response to high temperature

To assess the response of different P. xylostella strains (ancestral strain, hot strain and mutants) to extremely high temperatures, 2-day-old eggs, 1-day-old 3rd-instar larvae and 2-day-old pupae were individually placed in 90 mm diameter Petri dishes. Adult females and males were placed individually in perforated 1.5 mL centrifuge tubes. Based on a preliminary trial on the stage-specific temperature tolerance limit of P. xylostella (eggs, larvae, pupae, and both male and female adults of the ancestral and hot strains were placed in different temperature environments ranging from 40 to 45°C), pupae from the ancestral and mutant strains were exposed to 43°C while eggs, larvae, and adults were exposed to 42°C for periods ranging from 30 to 180 minutes. After treatment, all replicate samples were transferred to an environment maintained at 26°C, where survival was observed and recorded. Survival was defined as the successful development of eggs, larvae and pupae to the next stage, while adults had to show movement of an appendage or mouthparts. Experiments were performed with six biological replicates, with each replicate contained 20 individuals.

We randomly selected 20 female adults injected with dsRNA to determine their critical thermal maximum (CTMax). A thermistor probe (Omega, USA) was inserted into a 1.5 mL centrifuge tube, which was suspended inside a 50 mL centrifuge tube with the opening sealed with cotton. This assembly was then placed in a 2 L glass beaker containing 1000 mL water, with the beaker top sealed with insulating foam board. The entire setup was positioned on a thermostatically controlled magnetic stirrer, where the temperature inside the 1.5 mL tube was increased at a constant rate of 0.5°C/min. When the temperature reached 26°C, female adults were quickly transferred into the 1.5 mL centrifuge tube containing the temperature probe, and their behavioral responses were continuously monitored as temperature increased. The CTMax was recorded when moths exhibited spasms, lost their crawling or flying ability, and remained motionless at the bottom of the tube, typically lying ventral side up (in most cases) or dorsal side up (in fewer instances). Although antennae and limbs might still exhibit slight tremors at this point, the insects typically died within seconds (Li et al., 2024b).

Measurement of the supercooling and freezing points

To investigate the cold hardiness of different P. xylostella strains (including the ancestral, cold and mutant trains), we randomly selected 40 pupae from each strain to examine their supercooling and freezing points. A thermistor probe from a subcooling point tester (Omega, USA) was attached to a pupa, secured with conductive tape and placed in a centrifuge tube, with the tube mouth sealed by cotton. The centrifuge tubes were then placed in a 50 mL plastic cup filled with ethylene glycol (antifreeze), and the cup was stored in an ultra-low temperature freezer set at −70°C, with the temperature first dropping rapidly and then decreasing at a rate of 0.10°C per second until the supercooling point was reached. By recording temperature changes at intervals of every second, the supercooling and freezing points of pupae were determined based on the inflection point of body temperature. The same experimental approach was also applied to P. xylostella injected with dsRNA.

Detection of oxidative stress indicator

Samples were collected for the detection of superoxide dismutase (SOD) and superoxide anion (O2-) in the following conditions: (1) the eggs, 3rd-instar larvae, pupae, and adult males and females of the ancestral, hot, cold, and SODC-MU (MU1 and MU2) strains at the favorable temperature (26°C); (2) the 3rd-instar larvae of the ancestral, hot, and SODC-MU (MU1 and MU2) strains after 2 hours of heat stress at 32°C, 34°C, and 36°C; and (3) the 3rd-instar larvae of the ancestral, cold, and SODC-MU (MU1 and MU2) strains after 2 hours of cold stress at 12°C, 10°C, and 8°C. The experiment was performed with four independent biological replicates. The levels of SOD and O2- were measured using commercial assay kits (Comin, China) according to the manufacturer’s instructions.

Detection of 5-methylcytosine concentration

Pupae and female adults were collected for detection of 5-methylcytosine (5-mC) concentration after injection with dsRNA. The levels of 5-mC were measured using a commercial insect 5-methylcytosine (5-mC) ELISA detection kit (Shanghai Enzyme-linked Biotechnology Co., Ltd., China) according to the manufacturer’s instructions. The experiment was performed with four independent biological replicates.

Data analysis

Statistical analysis was performed using SPSS version 23.0. First, the Shapiro-Wilk test was used to assess the normality of data distribution. For the normally distributed data, independent samples t-tests were used to analyze the difference between the control and treated groups. For the data involving three or more groups with homogeneity of variances, one-way ANOVA followed by Tukey’s multiple comparison test was used, and if variances were uneven, Tamhane’s T2 test was used. For the non-normally distributed data, the Mann-Whitney test was used to analyze the difference between two groups, or the Kruskal-Wallis test was used to compare the differences among three or more groups (Lei et al., 2024).

eLife Assessment

This important study deepens our understanding of how populations of a given species may diverge in their molecular and physiological patterns as a result of adaptation to different thermal regimes. By approaching this question from multiple directions, the authors provide solid evidence for adaptive changes in three strains of the diamondback moth after only three years of experimental evolution, and support the causal involvement of the PxSODC gene in thermal adaptation to both cold and hot temperatures. This work would benefit from more sophisticated phylogenetic analyses, better statistical support, and a more detailed discussion of the differences in the three strains at the pathway level.

Reviewer #1 (Public review):

Summary:

In this manuscript, Lei and co-workers aim to uncover the genetic underpinnings of thermal adaptation across three strains of the diamondback moth (Plutella xylostella) through experimental evolution over three years under three different thermal regimes. They identify systematic differences in trait responses (e.g., survival, fecundity), metabolic profiles, gene expression, and in the amino acid sequence of the PxSODC gene, among others. These results suggest that the diamondback moth has a strong potential for rapid physiological adaptation to different thermal regimes. Overall, this is a comprehensive and generally well-executed study that addresses an important question in the face of ongoing climate change.

Strengths:

The authors employ multiple approaches to identify signatures of thermal adaptation across the three strains, such as trait performance comparisons, metabolomics, transcriptomics, and amino acid sequence comparisons. All these different angles form a convincing picture of the underlying factors that underpin thermal adaptation in this experimental system. The manuscript is also generally well written and easy to understand.

Weaknesses:

I am unable to judge the validity of all aspects of this work; I will focus only on areas within my core expertise.

(1) The authors identify pathways that are enriched in different strain comparisons (Figure 3E), but do not provide a detailed interpretation of these results. It would be great if the authors could explain in more detail how the physiological processes of a cold-adapted strain of this species may differ from those of a warmer-adapted strain.

(2) The authors reconstruct a phylogenetic tree of the PxSODC gene using the neighbor-joining algorithm. The limitations of this algorithm have been known for many years now, especially for sequences separated by long evolutionary distances. According to Wang et al. (2016), the last common ancestor of the species shown in Figure S4C occurred 392-350 million years ago. Given this, I would strongly recommend that the authors infer a phylogenetic tree using model-based methods, such as those implemented in RAxML-NG or IQ-TREE. Also, in the absence of a valid outgroup sequence, I would show the gene tree as unrooted or rooted based on the corresponding species tree.

(3) There is a key piece of the puzzle that is currently missing: the structural mechanism behind the mutational effects described in this study (e.g., Figure 5). The authors could leverage AlphaFold to generate structural models of different mutants and conduct molecular dynamics simulations to examine their conformational dynamics.

References:

Wang, Yh., Engel, M., Rafael, J. et al. Fossil record of stem groups employed in evaluating the chronogram of insects (Arthropoda: Hexapoda). Sci Rep 6, 38939 (2016). https://doi.org/10.1038/srep38939

Reviewer #2 (Public review):

Summary:

In this paper, the authors set out to better understand the genetic mechanisms underlying thermal adaptation in insects. They experimentally evolved diamondback moth (Plutella xylostella) populations - a pest species with a wide distribution - under both hot (12h:12h 32{degree sign}C/27{degree sign}C) and cold (15{degree sign}C/10{degree sign}C) thermal conditions, and conducted phenotypic assays and metabolic and transcriptomic profiling to analyze how populations changed to deal with this thermal stress compared to the nonevolved ancestral population (constant 26{degree sign}C). Phenotypic assays showed that evolved hot populations had increased survival at high temperatures (42-43{degree sign}C) while evolved cold populations had lower freezing points compared to the ancestral population. When measured at the constant 26{degree sign}C conditions, metabolic and transcriptomic profiles of 3rd instar larvae from the evolved population were distinctive from the ancestral population, with a set of overlapping metabolic and transcriptomic pathways that were significantly differentially expressed in both hot and cold evolved populations compared to the ancestral. The authors narrowed down this set of candidate genes further by focusing on genes with high expression levels overall, whose expression profile was correlated with differentially expressed metabolites, and that contained mutants in both hot and cold strains. From this set, they chose the PxSODC gene for further functional validation, as it has previously been shown to be involved in the response of insects to abiotic stress with its antioxidative role in cellular defense. At the constant 26{degree sign}C, this gene showed lower expression across development in evolved strains compared to the ancestral population, while it showed similar expression patterns under thermal stress. Knockdown of PxSODC resulted in decreased survival rates at high temperatures and higher freezing points compared to the ancestral population. Based on this validation, the authors hypothesize that the non-synonymous mutation in the PxSODC gene that they found in the cold and hot evolved populations might alter the conformation of the PxSODC protein, increasing enzyme capacity. Their experimental evolution experiment furthermore indicates the capacity of the pest species, the diamondback moth, to adapt to a wide range of temperatures, providing insights into its capacity for global dispersal.

Strengths:

(1) The authors did a tremendous amount of work to characterize the mechanisms underlying thermal adaptation in the diamondback moth, artificially selecting populations for three years in the lab and characterizing how they evolved as a result at different biological levels: from phenotypes in different life stages, to larval metabolites and gene transcription, to functionally validating how one of the resulting gene candidates influences the capacity to deal with thermal stress.

(2) The paper identifies and provides further evidence for candidate genetic mechanisms that might be particularly important for thermal adaptation in insects, including lipid metabolism, oxidoreductase activity, and DNA methylation. It is furthermore interesting that the authors found similar mechanisms to be involved in both the adaptation to cold and hot environments. Their functional validation of some of the genes involved in these mechanisms is very useful to understand how these genes might be causally involved in insect thermal adaptation.

(3) The paper also has applied value: the diamondback moth is a pest species with a wide distribution, so understanding its adaptive capacity to different thermal environments is important for predicting the prevalence and potential further range expansion of this species under future climate change.

Weaknesses:

(1) The paper in its current form is hard to digest and would benefit from improved clarification of the storyline, as well as a tighter integration between the phenotypic, omics, and functional validation data. Currently, it is not always clear what the relevance is of all the reported results, nor why certain decisions were made, or how all the different methods the authors used fit together. For example, the authors functionally validated a second gene, PxDnmt1, but it is unclear why this particular gene was chosen, nor how it relates to their selection regimes when looking at the results obtained with the phenotyping and omics data collection. Seeing how much work the authors did, this makes the paper overwhelming and difficult to read.

(2) The authors at times stretch their results too far, as the ecological relevance of their study design and results is not clear, limiting the generalizability and value of the results for understanding species' adaptive potential under climate change. For example, the selection regimes used present the minimum and maximum known temperatures at which the species can survive and develop, but it is unclear how the temperatures relate to the natural environment of the source population, to what extent wild populations might experience these temperatures, and whether they would experience them at the extended duration used (12h at max/min temperature). Moreover, I wonder whether the comparisons made would identify the genes that matter under natural conditions, as unevolved populations were kept under constant conditions compared to 12h:12h temperature regimes for the evolved populations, and the metabolic and transcriptomic profiling was done under a constant favorable 26{degree sign}C rather than under thermal stress in a, as far as I can tell, randomly chosen life stage (larval stage).

(3) The paper in its current form does not adequately describe the statistical analyses underlying the results, nor do the authors share their code, making it very hard to judge whether the analyses used are appropriate and the results trustworthy. I have concerns about the inappropriate use of t-tests, the lack of correcting for confounding variables, and the need for multiple testing corrections.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This important study deepens our understanding of how populations of a given species may diverge in their molecular and physiological patterns as a result of adaptation to different thermal regimes. By approaching this question from multiple directions, the authors provide solid evidence for adaptive changes in three strains of the diamondback moth after only three years of experimental evolution, and support the causal involvement of the PxSODC gene in thermal adaptation to both cold and hot temperatures. This work would benefit from more sophisticated phylogenetic analyses, better statistical support, and a more detailed discussion of the differences in the three strains at the pathway level.

We sincerely thank the editors for this positive and constructive assessment. In the revised manuscript, we have addressed the highlighted points by: (1) re-inferring the phylogenetic tree of the PxSODC gene using a model-based Maximum Likelihood method (IQ-TREE) to ensure a robust evolutionary analysis; (2) substantially expanding the description of our statistical methods across all data types to ensure reproducibility and clarify multiple-testing corrections; and (3) adding a more detailed discussion of the pathway-level differences between the hot and cold strains, particularly integrating how their distinct transcriptomic responses align with their shared metabolic adjustments and phenotypic traits.

Reviewer #1 (Public review):

(1) The authors identify pathways that are enriched in different strain comparisons (Figure 3E), but do not provide a detailed interpretation of these results. It would be great if the authors could explain in more detail how the physiological processes of a cold-adapted strain of this species may differ from those of a warmer-adapted strain.

We agree. We have addressed this by directly integrating our pathway enrichment results (Figure 3E) with the observed life-history phenotypes (concurrently addressing Reviewer 2's Comment 36a). We expanded the Discussion to explain that while both strains share convergent adjustments in core pathways (e.g., lipid metabolism for energy reallocation), their specific physiological strategies differ. The cold-adapted strain relies on broader transcriptional reprogramming to maintain homeostasis and support extended longevity/cold hardiness, whereas the hot-adapted strain utilizes broader metabolic rewiring to actively fuel its accelerated development and higher fecundity.

(2) The authors reconstruct a phylogenetic tree of the PxSODC gene using the neighbor-joining algorithm. The limitations of this algorithm have been known for many years now, especially for sequences separated by long evolutionary distances. According to Wang et al. (2016), the last common ancestor of the species shown in Figure S4C occurred 392-350 million years ago. Given this, I would strongly recommend that the authors infer a phylogenetic tree using model-based methods, such as those implemented in RAxML-NG or IQ-TREE. Also, in the absence of a valid outgroup sequence, I would show the gene tree as unrooted or rooted based on the corresponding species tree.

Agree. We have re-inferred the phylogenetic tree of the PxSODC gene using the model-based Maximum Likelihood (ML) method implemented in IQ-TREE. As recommended, in the absence of a valid outgroup sequence, the revised tree is now presented as unrooted. Supplemental Figure S4C (Figure 5-figure supplement 1C) and the corresponding text in the manuscript have been updated.

(3) There is a key piece of the puzzle that is currently missing: the structural mechanism behind the mutational effects described in this study (e.g., Figure 5). The authors could leverage AlphaFold to generate structural models of different mutants and conduct molecular dynamics simulations to examine their conformational dynamics.

We thank the reviewer for this excellent suggestion. We generated AlphaFold structural models of the wild-type (WT) and mutant (MU) PxSODC proteins and conducted 100 ns molecular dynamics (MD) simulations using GROMACS 2022.3 at three physiologically relevant temperatures: 15°C (cold stress), 26°C (favorable baseline), and 32°C (heat stress). Using 26°C as the physiological baseline, three key structural parameters support enhanced thermostability of the mutant protein (Figure 5–figure supplement 3). First, RMSD analysis revealed that under heat stress (32°C), the WT underwent severe conformational drift (RMSD increased from the 26°C baseline of 1.62 to 2.49, an increase of 0.87), while MU remained remarkably stable (from 1.59 to 1.66, an increase of only 0.07). Second, MU possessed a significantly more compact structure, with lower SASA values at 15°C (118.39 vs. 127.29 nm²) and 26°C (113.82 vs. 125.61 nm²), indicating optimized hydrophobic core packing. Third, the intramolecular hydrogen bond network of MU demonstrated dual stress resistance: under cold stress, MU actively increased hydrogen bonds from its baseline (113→119), whereas WT lost bonds (117→112); under heat stress, MU fully maintained its bond count (113→113). These results provide a direct structural mechanism for the enhanced catalytic efficiency of the mutant SOD at lower expression levels.

Reviewer #1 (Recommendations for the authors):

(4) The experimental evolution component of this study is described in the text as lasting for three years. It would help if the number of generations per strain were also reported.

We have added the number of generations per strain. Over the three-year period, the hot strain completed ~75 generations and the cold strain ~15 generations. The ancestral strain was continuously maintained at 26°C throughout this period. The revised text has been updated in both the Introduction and Materials and Methods.

(5) In Figure 3B: There is a typo in the word “Statistics”.

Corrected. The typo in “Statistics” in Figure 3B has been fixed.

(6) In Figure 3D: “CS” appears twice.

Corrected. The duplicated “CS” label in Figure 3D has been replaced with the correct label.

(7) Figure 4: This is not accessible to colorblind readers, who will clearly not be able to tell each color apart. As a non-colorblind person, I, too, have trouble figuring out which color label in panel B corresponds to which color in panel A. For example, I do not know off the top of my head how 'blue' differs from 'midnightblue', 'royalblue', or 'skyblue'. I recommend that the authors replace colors with identifiers, such as 'g1' for group 1 and so on.

We appreciate this suggestion. We have replaced all color-based module labels with alphanumeric identifiers (M1, M2, M3, etc.) and added a corresponding legend. The main text and supplementary materials have been updated accordingly.

(8) Lines 246-247: "Its secondary structure mainly consisted of strands, helices and coils." This sentence is redundant. These three are the only possible secondary structural elements, according to most bioinformatics tools such as PSIPRED, which the authors used. This sentence would be more useful if the authors could report the percentage breakdown of each secondary structural element.

We have removed the redundant sentence and updated the text to report the specific percentage breakdown of the secondary structural elements based on our PSIPRED predictions (approximately 55.24% random coils, 16.19% alpha helices, and 28.57% extended strands). The revised text has been updated in the Results section.

(9) Lines 260-261: "This suggests that the PxSODC gene can alter its expression pattern and function in response to environmental change...". I find this sentence a bit imprecise. Would it not be more precise to mention that the expression of this gene is regulated by temperature triggers?

We agree that the original phrasing was imprecise. We have revised the sentence in the manuscript to state: “This suggests that the expression of the PxSODC gene is regulated by temperature triggers, and its altered function contributes to temperature-adaptive evolution in P. xylostella.”

(10) The data points in Figures S1 and S7 are very small and hard to tell apart without zooming in a lot. Perhaps the authors could change the orientation of those pages to landscape and increase the size of the figures.

Done. We have changed the orientation of Supplemental Figures S1 (Figure 1-figure supplement 1) and S7 (Figure 5-figure supplement 4) to landscape and increased the size of the figures and individual data points to improve visibility.

(11) In Figure S2, the panel labeled as 'C' should be 'B' (based on the caption) and vice versa.

Corrected. The panel labels ‘B’ and ‘C’ in Supplemental Figure S2 (Figure 2-figure supplement 1) have been swapped. The Supplementary Materials have been updated accordingly.

Reviewer #2 (Public review):

(1) The paper in its current form is hard to digest and would benefit from improved clarification of the storyline, as well as a tighter integration between the phenotypic, omics, and functional validation data. Currently, it is not always clear what the relevance is of all the reported results, nor why certain decisions were made, or how all the different methods the authors used fit together. For example, the authors functionally validated a second gene, PxDnmt1, but it is unclear why this particular gene was chosen, nor how it relates to their selection regimes when looking at the results obtained with the phenotyping and omics data collection. Seeing how much work the authors did, this makes the paper overwhelming and difficult to read.

We sincerely appreciate this constructive feedback. In the revised manuscript, we have made significant structural revisions to improve the storyline and logical flow. We have streamlined the Results section (moving extensive descriptive data like life table curves and detailed metabolomics of mutant strains to the Appendix 1-3) to focus on the key findings. Furthermore, we have clarified the logical transitions between experiments. For instance, regarding the choice to validate PxDnmt1, we now explicitly explain in the Results that our untargeted metabolomic analysis of the PxSODC mutant strains revealed consistent alterations in 5-hydroxymethyluracil (involved in DNA demethylation) and 5'-deoxyadenosine (a precursor to the primary methyl donor S-adenosylmethionine) across all developmental stages. This specific metabolic signature provided a strong, data-driven hypothesis linking PxSODC function to epigenetic regulation via DNA methylation, prompting us to functionally validate PxDnmt1. By explicitly stating these rationales, the narrative is now much clearer and cohesive.

(2) The authors at times stretch their results too far, as the ecological relevance of their study design and results is not clear, limiting the generalizability and value of the results for understanding species' adaptive potential under climate change. For example, the selection regimes used present the minimum and maximum known temperatures at which the species can survive and develop, but it is unclear how the temperatures relate to the natural environment of the source population, to what extent wild populations might experience these temperatures, and whether they would experience them at the extended duration used (12h at max/min temperature). Moreover, I wonder whether the comparisons made would identify the genes that matter under natural conditions, as unevolved populations were kept under constant conditions compared to 12h:12h temperature regimes for the evolved populations, and the metabolic and transcriptomic profiling was done under a constant favorable 26°C rather than under thermal stress in a, as far as I can tell, randomly chosen life stage (larval stage).

We appreciate the reviewer raising these important points regarding ecological relevance and experimental design. In the revised manuscript, we have added context and acknowledged these limitations in the Methods and Discussion sections. First, regarding ecological relevance: The source population is from Fuzhou, a subtropical region where summer high temperatures frequently exceed 32°C and winter lows can drop below 10°C, making our selection temperatures ecologically relevant extremes for this population. The 12h:12h cycling temperatures were designed to simulate severe but natural diurnal fluctuations.

Second, regarding constant control vs. cycling regimes: The constant 26°C represents the established optimal developmental temperature and standard laboratory condition for P. xylostella. We acknowledge that comparing cycling selection regimes against a constant control might conflate adaptation to absolute temperature extremes with adaptation to thermal fluctuation itself. We have added this as a caveat in the Discussion. Third, regarding omics profiling conditions: The transcriptomic and metabolomic profiling was conducted under common garden conditions (26°C) specifically to identify constitutive, genetically fixed adaptations resulting from evolutionary selection, rather than immediate physiological plasticity under stress. We have clarified these rationales in the text.

(3) The paper in its current form does not adequately describe the statistical analyses underlying the results, nor do the authors share their code, making it very hard to judge whether the analyses used are appropriate and the results trustworthy. I have concerns about the inappropriate use of t-tests, the lack of correcting for confounding variables, and the need for multiple testing corrections.

We sincerely appreciate this concern. In the revised manuscript, we have made substantial improvements to the description of statistical analyses throughout the Methods section:

(1) Statistical methods for each data type are now described separately and in detail, specifying the tests used, the number and type of comparisons, and sample sizes.

(2) For metabolomic data, we have clarified that FDR correction was applied alongside multi-criteria thresholds (|log2Fold Change| ≥ 1, VIP ≥ 1, FDR < 0.05). For transcriptomic data, FDR correction (Benjamini and Hochberg, 1995) was applied via DESeq2.

(3) For WGCNA, we have specified the total number of correlation tests (29 modules × 30 metabolites = 870) and the stringent dual threshold (|r| > 0.8, P < 0.05) used to control for false positives, following standard practice.

(4) For life table parameters, the paired bootstrap method with 100,000 replications was used for all pairwise comparisons among strains.

(5) For all other experimental data (qRT-PCR, SOD activity, O2- levels, survival rates, supercooling/freezing points, etc.), we have specified that t-tests were used only for two-group comparisons, while one-way ANOVA with Tukey's or Tamhane's T2 test was used for three or more groups, with non-parametric alternatives applied when normality assumptions were not met.

(6) The raw data have been deposited in public repositories (see Data availability), and all statistical procedures are now described in sufficient detail to enable independent reproduction of the results.

Reviewer #2 (Recommendations for the authors):

Title

(4) I don't feel the title adequately captures the work, I would instead of 'adaptive evolution' use 'experimental evolution' and I would not use the word 'underpins' but instead 'indicates', as it is not clear from your work whether the adaptations to the lab conditions you used would be ecologically relevant nor whether they are involved in thermal adaptation in wild populations.

Accepted. The title has been revised to: “Experimental evolution to thermal stress indicates climate resilience in a cosmopolitan arthropod.”

Abstract

(5a) Please add the phenotype results to the abstract.

We have added key phenotype results to the abstract. The revised text now reads: “The hot strain showed accelerated development, higher fecundity, and increased survival under extreme heat, while the cold strain exhibited lower supercooling and freezing points, indicating enhanced cold hardiness.”

(6b) The Abstract doesn't really detail the answer to your research question yet: so what insights into the genetic mechanisms underlying thermal adaptation did you gain that are novel?

We agree. We have revised the Abstract to explicitly highlight the novel genetic and molecular mechanisms we discovered. Specifically, we now detail that thermal adaptation is driven by a coordinated mutational, metabolic, and epigenetic (1) an energy-efficient genetic mechanism where non-synonymous mutations in PxSODC enhance superoxide scavenging efficiency, enabling effective oxidative stress management at lower gene expression levels; (2) convergent metabolic adjustments, notably a reduction in lipid metabolism to conserve energy; and (3) epigenetic regulation of thermal tolerance via DNA methylation. The revised text has been updated in the Abstract accordingly.

(7c) Line 3: replace 'ectotherms' with 'arthropods' to match the title?

Done. “Terrestrial ectotherms” has been replaced with “terrestrial arthropods” in the abstract.

(8d) Line 9: replace 'demographic' with 'life history'?

Done. “Demographic” has been replaced with “life history” in the abstract.

Introduction

(9a) The storyline is a bit unclear. Do you want to focus on the increased threat from insect pests under climate change or on the threat of climate change on insect persistence? Please pick one and adapt your storyline accordingly. I would suggest focusing on the first and talking more about the range extension of pest species under climate change (which would also require adaptation to cold extremes).

We agree and have refocused the Introduction on the increased threat from insect pests under climate change, emphasizing that range expansion into new regions requires adaptation to both heat and cold extremes. Both the first and second paragraphs have been revised accordingly.

(10b) Line 31-33: What do you mean by 'shows a positive relationship between the thermal tolerance range and the level of climatic variability'? Are they able to tolerate a larger range of temperatures?

This sentence has been revised as part of the restructured Introduction, which now focuses on the range expansion of pest species under climate change. The revised text reads: “Such range expansion requires adaptation not only to warmer conditions in existing habitats but also to cold extremes encountered during colonization of higher latitudes or elevations (Harvey et al., 2020).”

(11c) Line 33-35: Is this information relevant here?

Agreed. This sentence has been removed as part of the restructured Introduction, which now focuses on the threat of pest range expansion under climate change.

(12d) Line 55-56: What exactly do we not know yet about the mechanisms that enable thermal adaptation that you aim to fill in this paper? Please rephrase your knowledge gap to be more concrete (e.g., "but we do not yet know how...").

We have rephrased the knowledge gap to be more concrete and aligned with the revised storyline. The revised text now reads: “...we do not yet know how long-term thermal selection drives coordinated changes across gene function, metabolic networks, and life history traits to enable thermal adaptation and range expansion in pest species.”

(13e) Line 57: Also, here, the storyline is unclear. Why did you use the diamondback moth as your model species? You provide many different reasons, but it would help if you emphasized one reason that is in line with whichever storyline you want to focus on: is it because it is an insect pest that can tolerate a wide range of temperatures?

We have streamlined this paragraph to focus on the primary rationale: P. xylostella is a globally distributed pest that thrives across a wide range of thermal environments, making it an ideal model for studying the genetic mechanisms of thermal adaptation. Supporting details on genomic resources are retained briefly as they enable the multi-omics approach used in this study.

(14f) Line 65: Demonstrated how? Please give a short summary of the evidence for their genetic capacity to tolerate future climates.

We have added a brief summary of the evidence. Specifically, genome-wide SNP analysis of field populations from 114 locations across diverse biogeographical zones revealed climate-adaptive genetic variability, indicating that P. xylostella can tolerate projected future climates in most regions (Chen et al., 2021).

(15g) Line 72: What does 'Age-stage' mean? Should it read 'Aged-staged'?

“Age-stage, two-sex life table” is an established demographic method developed by Chi (1988) that simultaneously accounts for both age and developmental stage in both sexes. This is a standard term in the field (Chi et al., 2020), so we have retained the original wording but added a brief clarification upon first use.

(16h) Line 78-80: This needs a bit more explanation. Why does an increased ability to scavenge superoxide anions affect adaptability under extreme temperature environments?

We have added a brief explanation. Extreme temperatures induce oxidative stress by elevating intracellular reactive oxygen species (ROS), including superoxide anions, which can damage cellular structures. Enhanced scavenging capacity thus helps maintain cellular homeostasis under thermal stress.

(i) Line 82-86: Please be more precise. What novel insights did you gain about the genetic mechanisms underlying thermal adaptation?

We have revised this sentence to more precisely summarize the novel insights, encompassing both the multi-omics findings and the functional validation of PxSODC.

Results

(18a) The results section is very long and presents an overload of information at the moment, overwhelming the reader. Consider moving some sections to the Supplements (for example, a large part of the phenotypic data that cannot be linked to the omics data and the metabolic profiling of the mutant strains) or leave them out of the paper altogether.

We agree that the Results section was too dense. We have streamlined it by moving the following content to the Supplementary Materials:

(1) Detailed age-stage survival and fecundity curve data for the ancestral, hot and cold strains (Supplementary Text S1).

(2) Detailed life table analysis of the PxSODC mutant strains (Supplementary Text S2).

(3) Detailed untargeted metabolomic profiling of the SODC-MU mutant strains across developmental stages (Supplementary Text S3).

The main text now retains only the key life history comparisons, extreme temperature tolerance results, omics-based evidence linking transcriptomics and metabolomics, functional validation of PxSODC, and the DNA methylation findings, with brief summaries and cross-references to the Supplements for supporting details.

(19b) Please also provide the effect sizes for the different effects you report, for example, how many degrees difference was there between ancestral and cold strains in the supercooling/freezing points, and what was the variation?

We have added specific effect sizes (mean ± SEM and between-group differences) for all key comparisons throughout the Results section, including preadult duration, stage-specific survival rates under extreme heat, supercooling/freezing points, and SODC-MU mutant strain comparisons. For example, the supercooling points of CS pupae (-23.99 ± 0.18°C) were 0.90°C lower than AS (-23.09 ± 0.26°C), and the freezing points were 2.66°C lower (-14.24 ± 0.61°C vs. -11.58 ± 0.52°C). Please refer to the revised manuscript for all updated values.

(20c) Line 93-94: "Intrinsic and finite rate of increase" of what?

Clarified. These are population growth parameters. The revised text now specifies “intrinsic rate of increase (r) and finite rate of increase (λ) of the population.”

(21d) Line 98-99: Please start the paragraph with this summary of the results and then further detail them.

We have restructured this paragraph by moving the summary sentence to the beginning, followed by the supporting details.

(22e) Line 100-109: Why did you look at daily survival and fecundity rates? Please add why this is relevant.

As part of the overall streamlining of the Results section, this paragraph on detailed age-stage survival and fecundity curves has been moved to Supplementary Text S1. A brief justification for their relevance has been added there, noting that these curves capture stage-specific variation in survival and fecundity that summary life table parameters alone may obscure.

(23f) Line 106: What do HS, AS, and CS stand for? And please provide the statistics for comparison of daily survival rates between the strains.

We have defined the abbreviations (HS = hot strain, AS = ancestral strain, CS = cold strain) at their first appearance in the Results section. This paragraph on daily survival and fecundity has been moved to Supplementary Text S1, where the abbreviations are also defined. The survival rates reported are the maximum daily survival rates derived from the age-stage specific survival rate curves (sxj), and the statistical comparisons among strains are presented in Supplemental Table S1.

(24g) Line 144-146: Why are these differential metabolites likely to play a crucial role?

We agree this statement was speculative. It has been removed from the revised manuscript.

(25h) Line 159-161: Why is a reduction of lipid metabolites evidence for adaptive evolution?

We have revised this sentence to clarify the reasoning. The reduction in lipid metabolites in both independently evolved hot and cold strains suggests a convergent metabolic response, indicating that lipid metabolism adjustment is a shared adaptive strategy rather than a random change.

(26i) Line 184-185: It is difficult to judge from Figure 3E the extent of overlap in KEGG pathways between the hot and cold strains. Can you adjust the figure to emphasize that overlap more?

Agree. To intuitively emphasize the extent of overlap in KEGG pathways between the hot and cold strains, we have completely redesigned Figure 3E. Instead of presenting two separate panels with unaligned vertical axes, we have consolidated the data into a single back-to-back (mirrored) bar chart with a shared central y-axis.

(27j) Line 211: Not only the red module, but also the blue and green module correlates with many of the shared differential metabolites.

We agree. We have revised the text to acknowledge that the blue and green modules also showed strong correlations with shared differential metabolites, while noting that the red module had the highest number of significantly correlated metabolites and was therefore selected for further analysis.

(28k) Line 215: I would rephrase this as genes being interesting candidates for being involved in thermal adaptation or 'seem to be important for the adaptation of...', as you don't know from these results whether these genes play a critical regulatory role.

Agreed. We have toned down the language to reflect the correlative nature of these results.

(29l) Line 233: Do you mean that you further analyzed 15 genes of the 79 identified candidate genes in the previous paragraph?

Yes, exactly. From the 79 candidate genes, we selected 15 that were both annotated in the genome and had high expression levels (FPKM > 10) for further analysis. We have clarified this in the revised manuscript.

(30m) Line 238: What does SOD stand for?

We have spelled out the abbreviation upon first use in this section.

(31n) Line 254-255: Please provide the stats for this result.

We have added the specific allele frequencies for each strain. The Leu194-Met194 mutation frequency was determined by direct sequencing of 10 individuals per strain, and the frequencies are now reported in the revised text.

(32o) Line 303-304: How did you test for enhanced stability to temperature fluctuations? And enhanced compared to what?

This observation was based on the survival rate data in Figure 5C, where mutant pupae at 43°C showed no significant difference from the ancestral strain, whereas other life stages (eggs, larvae, adults) at 42°C showed significantly reduced survival in the mutant strains. We have revised the text to clarify the comparison.

(33p) Line 324-326: Why do decreased expression levels demonstrate increased O₂⁻ scavenging capacity? And why is that beneficial for adaptation to thermal stress? Please explain.

We have revised this sentence to clarify the logic. The non-synonymous mutations in the hot and cold strains likely alter the protein conformation of SOD enzymes, increasing their catalytic efficiency per molecule. This allows effective O2- scavenging at lower expression levels, which is energetically favorable under thermal stress where energy conservation is critical for survival.

(34q) Line 404-406: I'm confused. Is there a direct link between the gene you knocked out here and the results you presented up until now? How do the reduced levels of 5-methylcytosine relate to the metabolite results you present at the beginning of the paragraph, other than that both could be involved in DNA methylation?

We have revised this paragraph to clarify the logical chain. Among the three metabolites consistently altered across all developmental stages in the SODC-MU strains, 5-hydroxymethyluracil is involved in dynamic DNA demethylation and 5'-deoxyadenosine is a precursor to S-adenosylmethionine (the methyl donor for DNA methylation). This suggested a link between PxSODC deletion and DNA methylation. To test this, we examined PxDnmt1 expression and activity in the thermally adapted strains and found both were significantly reduced. We then used RNAi to silence PxDnmt1 and confirmed that reduced DNA methylation (lower 5-mC levels) directly impaired thermal tolerance. Thus the connection is: PxSODC deletion → altered methylation-related metabolites → reduced DNA methyltransferase activity → decreased thermal tolerance.

(35r) Line 410: Saying that your knockdown of a gene that did not directly pop up in any of your other analyses confirms that DNA methyltransferase is associated with the response to thermal selection is a stretch. Please rephrase.

We agree this was overstated. We have toned down the language to reflect that the RNAi results provide preliminary evidence for a potential role of DNA methylation in thermal tolerance, rather than confirmation.

Discussion

(36a) The phenotype data are currently not discussed at all. Please add it to the discussion and try to integrate it more with the omics data you collected.

We agree. To provide a cohesive narrative and avoid redundancy, we have addressed this comment in conjunction with our pathway interpretation (please see our response to Reviewer 1, Comment 1). In the revised Discussion, we explicitly integrated our specific phenotypic findings (e.g., accelerated development, increased fecundity, and heat survival in the hot strain; prolonged lifespan and lowered supercooling points in the cold strain) with the distinct transcriptomic and metabolomic profiles. This integration demonstrates how molecular and metabolic rewiring directly underpins the divergent life-history traits without engaging in unwarranted speculation.

(37b) Line 433-434: I don't think this adequately represents the relevance of your particular study. I would suggest changing it to be more in line with the storyline of understanding the capacity for global dispersal in insect pests under climate change.

We agree. We have revised this sentence to align with the storyline of pest range expansion under climate change.

(38c) Line 476: This is a very odd statement; don't all species' genomes have genes encoding proteins involved in thermal adaptation? The reference also doesn't seem to be appropriate. I would suggest deleting this sentence.

Agreed. This sentence has been removed.

(39d) Line 483: Please write out SOD the first time you use it in a new section.

Done. SOD has been spelled out at its first use in the Discussion.

(40e) Line 544-548: This is a bit too specific to be the last sentence of the discussion. Try to formulate it more broadly in terms of what future research should focus on in general, not just your specific research.

We agree. We have broadened the final sentence to address future research directions more generally.

Figures

(41a) Figure 1A: I don't think t-tests are appropriate here since you are not simply comparing two treatments, but testing for the effects of 5-6 different temperatures. And how did you correct for replicate populations in your analysis?

Clarified. In Figure 1A, our comparisons are independent pairwise tests between exactly two strains (HS vs. AS) at each specific temperature and time point, making t-tests statistically appropriate. We were not testing for a continuous effect across temperatures. Regarding replicate populations, the individuals used in these assays were drawn from across the six replicate populations per treatment, with each biological replicate (n = 6, with 20 individuals per replicate) comprising individuals pooled from across the replicate populations to account for inter-population variation. We have clarified this in the revised figure legend.

(42b) Figure 1B, Figure 5D, Figure 7: bar graphs are used for count data, so do the data represent the number of individuals with a certain trait value? If they are instead showing the mean of the population/treatment group, please use mean points ± standard errors instead.

Accepted. The data in these figures represent continuous physiological traits (e.g., supercooling/freezing points) showing the mean of the populations, rather than count data. To align with current data visualization standards for continuous variables and to provide full transparency of the underlying data distribution, we have replaced the bar graphs in Figures 1B, 5D, and 7 with scatter plots. These revised figures now display the mean ± SEM overlaid with all individual biological replicate data points.

(43c) Figure 3B: There is a typo in the graph, it reads 'Stattistics' instead of 'Statistics'.

Corrected. The typo ‘Stattistics’ in Figure 3B has been fixed.

(44d) Figure 3C: I don't understand what the colors of the graph mean here. Is it the average differential expression of each replicate compared to the ancestral?

Clarified. We have updated the figure legend to explain that the colors represent the Pearson correlation coefficient (r) between pairs of biological replicates, indicating the degree of transcriptomic similarity among samples.

Methods

(45a) Please start each new methods paragraph with the purpose of the method/analysis, for example, "To investigate XX, we used method X to measure X". It is at the moment hard to understand why certain things were done.

We agree. We have revised each Methods paragraph to begin with a clear statement of purpose, so that the rationale for each analysis is immediately apparent. All changes are shown in the revised manuscript.

(46b) Line 575-578: Why were the selection regimes with cycling temperatures and the control with constant?

The cycling temperatures in the hot (32°C/27°C) and cold (15°C/10°C) regimes were designed to simulate diurnal temperature fluctuations (12h light/12h dark) that more closely reflect natural thermal environments. The control was maintained at a constant 26°C, which is the established optimal developmental temperature for P. xylostella (Liu et al., 2002) and represents the standard laboratory rearing condition. We acknowledge this asymmetry and have added a justification in the revised manuscript.

(47c) Line 581: How many generations was the ancestral population kept in the lab before the start of the selection experiment? And for how many generations were the populations selected?

The ancestral population was maintained in the laboratory for approximately ~170 generations (from July 2012 to the start of the selection experiment) before the thermal selection began. The hot strain was selected for ~75 generations and the cold strain for ~15 generations over the three-year experiment. We have added this information to the revised manuscript.

(48d) Line 585-586: I don't understand what you mean by randomly selecting six replicate populations per treatment for downstream experiments when you only had six replicate populations per treatment to begin with (as detailed in Line 574)?

We apologize for the confusion. All six replicate populations per treatment were used for downstream experiments. We have corrected this sentence to remove the misleading “randomly selected” wording.

(49e) Line 590: Were these 90 eggs also randomly selected, like for the individual life tables? And were these kept at the baseline temperature conditions?

Yes, the 90 eggs were randomly selected and maintained under the baseline favorable temperature (26°C). We have clarified this in the revised manuscript.

(50f) Line 606: Which life history and population fitness parameters were calculated?

We have specified all parameters calculated in the revised manuscript.

(51g) Line 609: Link to software doesn't work.

We have updated the software link to the current working URL.

(52h) Line 611: Please spell out what 'BT' stands for.

Done. “BT” has been spelled out as “bootstrap” upon first use.

(53i) Line 612-613: How many tests did you do? Did you correct for multiple testing? Using what method?

The paired bootstrap method implemented in TWOSEX-MSChart inherently accounts for multiple pairwise comparisons through 100,000 bootstrap replications. We have clarified the scope of comparisons in the revised manuscript.

(54j) Line 620-621: What does biological replicate mean here? Individual eggs / larvae / pupae / adults, or were all or some life stages pooled? Also, you now only detailed which samples were collected for metabolomic profiling, were the same samples used for transcriptomic profiling, or a subset?

Each biological replicate consisted of pooled individuals at the same developmental stage. The same sample collection strategy was used for both metabolomic and transcriptomic profiling, but from independent biological replicates (six for metabolomics, three for transcriptomics). We have clarified this in the revised manuscript.

(55k) Line 637: Also here, how many tests did you do? Were p-values corrected for multiple testing? Using what method?

Differential metabolites were identified through pairwise comparisons using Student's t-test with FDR correction for multiple testing. A multi-criteria threshold of |log2Fold Change| ≥ 1, VIP ≥ 1, and FDR < 0.05 was applied. This approach was used for all metabolomic comparisons, including HS vs. AS, CS vs. AS, and SODC-MU vs. AS. We have clarified this in the revised manuscript.

(56l) Line 662: And here: how many tests did you do? Did you correct for multiple testing? Using what method?

In the WGCNA analysis, Pearson correlations were calculated between each module eigengene and each of the 30 common differential metabolites, resulting in a total of 29 × 30 = 870 correlation tests. Following standard WGCNA practice, rather than applying FDR correction, we used a stringent dual threshold of |correlation coefficient| > 0.8 and P < 0.05 to identify significant module-metabolite associations, which effectively controls for false positives (Langfelder and Horvath, 2008). We have clarified this in the revised manuscript.

(57m) Line 663: How did you select these modules? The ones that significantly correlated with differential metabolites? Why did you not use the phenotype data here?

Modules were selected based on significant correlations (|correlation coefficient| > 0.8, P < 0.05) with differential metabolites shared between the hot and cold strains. We chose metabolites rather than phenotype data as the trait input for WGCNA because metabolites serve as intermediate molecular phenotypes that bridge gene expression and organismal phenotypes, providing a more direct link to the underlying regulatory mechanisms. This approach allowed us to identify gene modules most closely associated with the metabolic changes driven by thermal adaptation, which could then be connected to the observed life history and fitness divergence.

(58n) Line 666: move RNA extraction details to before RNAseq methods description.

Done. The “RNA extraction and cDNA synthesis” section has been relocated to before the “Transcriptomic profiling” section for better logical flow.

(59o) Line 836: This paragraph describing the statistics is very short, and it is unclear to what data the described analyses apply. As the different types of data are very different, I expect the analyses to differ as well. Please describe the statistical analyses for each data type in more detail, specifying what tests you used, which, and how many comparisons were performed.

We agree. The statistical methods for life table analysis, metabolomics, and transcriptomics have been detailed in their respective method sections. We have expanded the Data analysis section to specify the statistical tests for the remaining experimental data.

(60p) Line 837: Please include your SPSS scripts to ensure the reproducibility of your results.

The statistical analyses in SPSS were performed using the graphical user interface. As all statistical tests, parameters, and comparison groups have been described in detail in the revised Methods section, and the raw data have been deposited in public repositories (see Data availability), we believe the analyses are fully reproducible. We are happy to provide additional details if needed.

Version 2

Abstract

Adaptive evolution enables species to survive and thrive under changing environmental conditions. In the face of accelerating global climate change, thermal stress represents a major challenge to the persistence of terrestrial arthropods. Understanding the genetic mechanisms underlying thermal adaptation is therefore critical for predicting species’ evolutionary potential and future success. Here, we combine experimental evolution, phenotypic assays, and multi-omics analyses to investigate the adaptive responses of the diamondback moth (Plutella xylostella), a globally destructive pest of cruciferous crops, to contrasting thermal environments. Populations evolved under hot (32°C/27°C) and cold (15°C/10°C) regimes exhibited distinct life history and fitness traits relative to those maintained under favorable conditions (26°C). The hot strain showed accelerated development, higher fecundity, and increased survival under extreme heat, while the cold strain exhibited lower supercooling and freezing points, indicating enhanced cold hardiness. Integrated transcriptomic and metabolomic analyses revealed extensive transcriptional reprogramming and convergent metabolic adjustments, notably a reduction in lipid metabolism to conserve energy under thermal stress. Crucially, non-synonymous mutations in PxSODC enhance superoxide scavenging efficiency, enabling effective oxidative stress management at lower gene expression levels. Furthermore, we identified epigenetic regulation via DNA methylation as a key mediator of this thermal tolerance. Together, these coordinated mutational, epigenetic, and metabolic insights highlight this arthropod’s capacity for global dispersal and likely persistence under climate change, establishing a framework for understanding equivalent effects in other species.

Introduction

Human-induced climate change, particularly the continued change in temperature and precipitation patterns (IPCC, 2023), is altering the geographical distribution of insect pests, allowing those previously confined by temperature barriers to spread to new areas and posing growing threats to crop production and food security (Deutsch et al., 2008; Outhwaite et al., 2022; Lawlor et al., 2024). Such range expansion requires adaptation not only to warmer conditions in existing habitats but also to cold extremes encountered during colonization of higher latitudes or elevations (Harvey et al., 2020). As poikilothermic organisms with high surface area to volume ratios, insects are particularly responsive to temperature change (Wang et al., 2022), and species with higher genetic diversity and adaptive potential may be better equipped to cope with novel thermal environments, giving them an advantage in expanding into new habitats.

Adaptive evolution is a crucial mechanism for insect pests to expand their geographical ranges under climate change (McCulloch and Waters, 2022; Burc et al., 2025). Understanding the genetic basis of thermal adaptation is therefore essential for predicting how and ultimately where pest species will colonize new regions as temperature barriers shift (Gibert et al., 2019). Genetic mutations, particularly non-synonymous mutations, can alter protein structure or function and increase thermal tolerance (Belfield et al., 2018). For example, nonsynonymous mutations in the AcVIAAT gene of the eastern honeybee, Apis cerana (e.g., P42L substitution), are associated with enhanced thermal adaptation (Li et al., 2024c). and the alpine ground beetle, Nebria vandykei, achieve survival in extreme thermal environments through adaptive selection of mutations in key genes (TREH, EIF3A, LRPPRC, etc.) coupled with immediate responses from heat shock proteins (Schoville et al., 2024). Although significant progress has been made in identifying such mutations in non-model organisms, we do not yet know how long-term thermal selection drives coordinated changes across gene function, metabolic networks, and life history traits to enable thermal adaptation and range expansion in pest species.

The diamondback moth (DBM), Plutella xylostella (Lepidoptera: Plutellidae), is a globally distributed pest of cruciferous crops thriving across a wide range of climatic conditions (Furlong et al., 2013). Genome-wide SNP analysis of field populations from 114 locations revealed climate-adaptive genetic variability, suggesting that P. xylostella can tolerate projected future climates in most regions (Chen et al., 2021). These features, together with the availability of complete genome sequences and extensive SNP datasets (You et al., 2013; You et al., 2020), make it an ideal model for studying the genetic basis of thermal adaptation through integrated multi-omics approaches. Here, we investigate the mechanisms of P. xylostella’s genetic adaptation and evolutionary responses to different thermal environments. Specifically, we employ thermal regime patterns of 12h/12h in hot (32°C/27°C) or cold (15°C/10°C) environments over the course of three years (∼75 and ∼15 generations for the hot and cold strains, respectively), compared to the favorable constant condition at 26°C, to investigate the adaptive evolution of P. xylostella in climate-controlled chambers.

Age-stage, two sex life tables (a demographic method that simultaneously incorporates age, developmental stage, and both sexes; Chi, 1988) of P. xylostella measured the life history variation of the three P. xylostella strains evolved in the favorable (ancestral), hot and cold environments. Using metabolomic and transcriptomic analyses, we identified the key genes that could facilitate the adaptation of P. xylostella to thermal extremes. Our results showed that a large number of differentially expressed genes and metabolites were produced in populations adapted to high and low temperatures through multi-generational selection. We find the mutant of a key gene, PxSODC, which can alter the superoxide dismutase activity and increase the ability to scavenge superoxide anions. Since extreme temperatures elevate intracellular reactive oxygen species that damage cellular structures, this enhanced scavenging capacity helps maintain cellular homeostasis, thereby significantly affecting the adaptability under high and low temperature environments. CRISPR-Cas9 was used to functionally validate the role of PxSODC in facilitating adaptive evolution and influencing regulatory networks. These findings demonstrate that long-term thermal selection drives coordinated transcriptomic, metabolic, and life history divergence in P. xylostella, and identify non-synonymous mutations in PxSODC that enhance superoxide scavenging efficiency as a key genetic mechanism underlying thermal adaptation, providing a framework for predicting its population dynamics under global climate change.

Results

Life history trait divergence among temperature-adapted strains

Following three years of evolution under contrasting thermal regimes (∼75 and ∼15 generations for the hot and cold strains, respectively), the hot strain (HS), cold strain (CS), and ancestral strain (AS) exhibited divergent life history traits. The hot strain exhibited an accelerated life cycle and increased fecundity, while the cold strain had extended male longevity. Specifically, both hot and cold strains had a significantly shorter preadult duration than the ancestral strain. The hot strain also had significantly shorter female longevity, and oviposition days, while the cold strain had a longer male longevity when compared to the ancestral strain (Supplementary File 1). The female fecundity, population intrinsic rate of increase (r) and finite rate of increase (λ) were all significantly higher in the hot strain than the cold strain which was not significantly different than the ancestral strain (Supplementary File 1). Detailed age-stage survival and fecundity curves are provided in Appendix 1.

To assess the evolved thermal tolerance of the temperature-adapted strains, we further examined the stage-specific survival rates of the hot and ancestral strains under extremely high temperatures, as well as the supercooling and freezing points of the cold and ancestral strains at pupae stage. The survival rates of eggs, 3rd-instar larvae and adults in the hot strain were higher than those of the ancestral strain at 42°C (e.g., 3rd-instar larvae at 120 min: HS 26.67% ± 3.57% vs. AS 13.33% ± 2.47%), and the survival rate of pupae in the hot strain was higher than that in the ancestral strain at 43°C and 44°C (Figure 1A). The supercooling and freezing points of pupae in the cold strain (supercooling: -23.99 ± 0.18°C; freezing: -14.24 ± 0.61°C) were significantly lower than those in the ancestral strain (supercooling: -23.09 ± 0.26°C; freezing: -11.58 ± 0.52°C), with differences of 0.90°C and 2.66°C, respectively (Figure 1B). The variation in survival rates and the supercooling/freezing points at extreme temperatures suggest that the hot and cold strains of P. xylostella have undergone profound adaptive adjustments.

Evolved phenotypic changes in temperature-adapted strains.

(A) Stage-specific thermal tolerance responses (survival rates) of the ancestral and hot strains, with 20 individuals used in each of the six replicates for every treatment. (B) Supercooling and freezing points of pupae for the ancestral and cold strains, with 40 biologically independent samples used in each treatment. Data are presented as mean ± SEM. Statistical analyses are performed using t-tests with significant levels indicated by asterisks (*p < 0.05, **p < 0.01, ***p < 0.001).

Omics-based evidence for adaptive evolution

Our previous studies have identified metabolites such as trehalose and very long chain fatty acids that play a role in adaptation of P. xylostella to both high and low temperatures (Zhou et al., 2022; Lei et al., 2023). We performed a broad analysis of targeted metabolites of 3rd-instar larvae of each strain using high-throughput UPLC-MS/MS. A total of 781 metabolites were identified, including 199 amino acids and their metabolites, 146 lipids, 90 organic acids and their derivatives, 78 nucleotides and their metabolites, 61 heterocyclic compounds, 45 benzene and substituted derivatives, 42 alcohols and amines, 37 carboxylic acids and derivatives, 21 coenzymes and vitamins, and 62 other metabolites (Figure 2A). Principal component analysis (PCA) and inter-sample correlation heat maps revealed significant metabolic changes in the 3rd-instar larvae from the ancestral strain to the hot and cold strains (Figure 2B; Figure 2–figure supplement 1A). These comprised 77 differential metabolites (34 up-regulated, 43 down-regulated compared to the ancestral strain) in the hot strain, and 37 differential metabolites (13 up-regulated, 24 down-regulated compared to the ancestral strain) in the cold strain (Figure 2C).

Metabolomic analysis of 3rd-instar larvae across ancestral, hot and cold strains (AS, HS and CS).

(A) Classification of metabolites, with a total of 781 metabolites being identified in different strains. (B) Principal component analysis (PCA) of the 781 metabolites across different strains. PC1 and PC2 represent the first and second principal components, respectively. (C) Volcano plot showing the down-regulated (green dots) and up-regulated (red dots) metabolites based on comparison between HS/CS and AS. (D) Classification of differential metabolites between HS/CS and AS. (E) Venn diagram showing the common and unique differential metabolites in HS and CS as compared to AS. (F) Fold changes and classifications of the common differential metabolites in HS and CS as compared to AS.

Compared to the ancestral strain, the common differential metabolites of the hot and cold strains included lipids, amino acids and their metabolites, organic acids and their derivatives, nucleotides and their metabolites, and benzene and substituted derivatives (Figure 2D; Figure 2–figure supplement 1B). Inter-replicate analysis of differential metabolites showed a low correlation between the ancestral strain and hot/cold strains, but a high correlation between the hot and cold strains (Figure 2–figure supplement 1C). Notably, 30 common metabolites were identified across the differential sets based on comparison of the hot/cold strains to the ancestral strain (Figure 2E). These metabolites, except for N1, N8-diacetylpiperidine, exhibited similar fold changes when comparing the hot/cold strains to the ancestral strain (Figure 2F). These results indicate that P. xylostella responds to different environmental stresses by regulating similar metabolic pathways. Further analysis revealed a reduction in most of the lipid metabolites in both HS and CS compared to AS (Figure 2F).

We then profiled and compared the transcriptomes of the three strains to identify the key genes involved in adaptation of P. xylostella to temperature extremes. This revealed significant variation in gene expression among strains (Figure 3A; Figure 3–figure supplement 1A), with 1364 (825 up-regulated, 539 down-regulated) and 2029 (1205 up-regulated, 824 down-regulated) genes differentially expressed in the hot and cold strains, compared to the ancestral strain (Figure 3B). Pearson correlation showed, in contrast to the metabolomics data, a lack of strong correlation between the differentially expressed genes of the hot/cold strains and the ancestral strain (Figure 3C), with 498 common differentially expressed genes (Figure 3D). However, KEGG analysis revealed that the differentially expressed genes between the hot/cold strains and the ancestral strain were enriched in a substantial overlap of similar pathways, such as transport and catabolism, signal transduction, and lipid metabolism (Figure 3E), indicating that while multiple genes are involved in the adaptation of P. xylostella to high and low temperatures, a relatively limited range of biological functions might be affected.

Transcriptomic analysis of the 3rd-instar larvae across the ancestral, hot and cold strains.

(A) Principal component analysis (PCA) of genes across different strains. PC1 and PC2 represent the first and second principal components, respectively. (B) Volcano plots of differential gene expression, showing significantly up-regulated (red dots) and down-regulated (green dots) genes between HS/CS and AS (FDR < 0.05, fold change > 2). (C) Cluster analysis of the transcriptome. The colors represent the Pearson correlation coefficients between samples, indicating transcriptomic similarity. (D) The number of common or unique differentially expressed genes between HS/CS and AS. (E) KEGG function classification of differentially expressed genes between HS/CS and AS.

The gene expression-based clustering tree using a weighted gene co-expression network analysis (WGCNA) was divided into 29 modules as shown with alphanumeric identifiers (M1-M29) (Figure 4A; Figure 3–figure supplement 1B). Module M4 contained the most genes (3463), while module M29 contained the fewest (31) (Figure 4B). Selecting the common differential metabolites (30 in total) shared between the hot and cold strains as compared to the ancestral strain, we performed a correlation analysis with co-expressed networks and found that multiple modules, including modules M13, M19, and M24, showed strong correlations with shared differential metabolites. Module M13 was selected for further analysis as it had the highest number of significantly correlated metabolites (28 of 30) (Figure 4C; Figure 3–figure supplement 1C). Further analysis revealed that 79 genes within module M13 were differentially expressed in the hot and cold strains when compared with the ancestral strain (Figure 4D). These results suggest that genes in module M13 may be candidates involved in the adaptation of P. xylostella to extreme temperatures.

Weighted gene co-expression network analysis (WGCNA) of transcriptomes for the 3rd-instar larvae across the ancestral, hot and cold strains (HS, CS and AS) of P. xylostella.

(A) Hierarchical cluster tree illustrating 29 modules identified by WGCNA. (B) Numerical distribution of genes of different modules as identified by WGCNA clustering. (C) In WGCNA, module M13 shows the highest number of metabolites strongly correlated with genes. (D) Overlap of genes in module M13 from WGCNA with common differentially expressed genes between HS/CS and AS.

Genetic basis of temperature adaptation

To further elucidate the genetic basis of P. xylostella adaptation to hot and cold environments, from the 79 candidate genes identified above, we selected 15 that were annotated in the genome and had high expression levels (FPKM > 10) for further analysis (Figure 4D) and identified 11 genes being successfully amplified. Comparative results revealed eight genes with nonsynonymous mutations and one with a synonymous mutation in both the hot and cold strains (Supplementary File 2). Among these genes, we focused on the role of Px04C00666 (PxSODC) in temperature adaptation of P. xylostella because the deletion of superoxide dismutase (SOD) genes can alter the response of insects to abiotic stresses including temperature (Bittner et al., 2019; Quan et al., 2024).

The NCBI database predicted that PxSODC contained three exons and two introns, with a conserved domain belonging to the copper-zinc superoxide dismutase superfamily (Figure 5–figure supplement 1A) which plays an antioxidative role in cellular defense systems, protecting cells from damage caused by reactive oxygen species (Fridovich, 1995). The open reading frame of PxSODC was 633 bp and encodes 210 amino acids. Expasy predicted that the molecular weight of the PxSODC protein was 22,168.24 Da, with the isoelectric point being 6.29. According to PSIPRED predictions, its secondary structure consisted of approximately 55.24% random coils, 16.19% alpha helices, and 28.57% extended strands (Figure 5–figure supplement 1B). Evolutionary analysis using a Maximum Likelihood approach showed PxSODC of P. xylostella clustered with that of other Lepidoptera insects such as Operophtera brumata, Vanessa cardui, and Cydia fagiglandana, indicating its conserved evolution within this taxonomic order (Figure 5–figure supplement 1C). The coding region of PxSODC in the hot and cold strains had 23 SNP sites, including 20 synonymous and three non-synonymous mutations (Leu9-Val9, Lys25-Gln25, Leu194-Met194) (Figure 5A-B). Leu9-Val9 and Leu194-Met194 mutations were involved in the substitution of hydrophobic amino acids. Based on sequencing of 10 individuals per strain, the Leu194-Met194 mutation was present at a frequency of 70% in HS, 90% in CS, and 30% in AS (Figure 5B). The expression of PxSODC at different developmental stages of the hot and cold strains was significantly lower than that of the ancestral strain (Figure 5–figure supplement 2A). After 2 h exposure of the 3rd-instar larvae to the stress of high (32°, 34°, 36°, 38° and 40°C) or low (12°, 10°, 8°, 6° and 4°C) temperature environments, the expression of PxSODC in the hot and cold strains was significantly lower than in the ancestral strain (Figure 5–figure supplement 2B-C). This suggests that the expression of the PxSODC gene is regulated by temperature triggers, and its altered function contributes to the temperature-adaptive evolution in P. xylostella.

Role of PxSODC in temperature adaptation of P. xylostella.

(A) Allele frequencies of SNPs in the PxSODC gene amplified by PCR from the ancestral, hot and cold strains (AS, HS and CS). The analysis involves ten 4th-instar larvae from each of the strains; the dot (·) indicates identity with the reference base. (B) Frequency of amino acid translations from non-synonymous codon mutations in the PxSODC gene in different strains. (C) Stage-specific survival rates of the ancestral and mutant strains (AS, SODC-MU1, SODC-MU2 and SODC-MU3) under extreme heat conditions. (D) Supercooling and freezing points of the pupae from different strains (AS, SODC-MU1, SODC-MU2 and SODC-MU3). Data are presented as mean±SEM, one-way ANOVA with Tukey’s test was used for comparison.Six biologically independent samples were used in (C) and significant levels between groups with the same stress duration are indicated by asterisks (*p < 0.05, **p < 0.01, ***p < 0.001). A total of 40 biologically independent samples were used in (D) and statistical significance is indicated by different letters (p < 0.05).

To elucidate the structural mechanism by which these non-synonymous mutations enhance protein function under thermal stress, we performed 100 ns MD simulations on AlphaFold-generated models of the wild-type (WT) and mutant (MU) PxSODC at 15°C, 26°C (favorable baseline), and 32°C (Figure 5–figure supplement 3). RMSD analysis revealed that at the 26°C baseline, both WT and MU exhibited comparable structural stability (1.62 ± 0.21 vs. 1.59 ± 0.27) (Figure 5–figure supplement 3B). However, under heat stress (32°C), WT underwent severe conformational drift (RMSD surged to 2.49 ± 0.35, an increase of 0.87 from baseline), while MU remained remarkably stable (1.66 ± 0.26, an increase of only 0.07) (Figure 5–figure supplement 3C). SASA analysis showed that MU possessed an inherently more compact structure than WT, with lower values at both 15°C (118.39 ± 7.57 vs. 127.29 ± 6.12 nm2) and 26°C (113.82 ± 7.40 vs. 125.61 ± 6.76 nm2), indicating optimized hydrophobic core packing (Figure 5–figure supplement 3D). Analysis of intramolecular hydrogen bonds further revealed that the MU network exhibited dual stress resistance: under cold stress (15°C), MU actively increased hydrogen bonds from its 26°C baseline (113→119), demonstrating a compensatory response, while WT suffered bond loss (117→112); under heat stress (32°C), MU fully maintained its hydrogen bond count (113→113), whereas WT showed a slight decrease (117→116) (Figure 5–figure supplement 3E). Collectively, these simulations demonstrate that the non-synonymous mutations confer enhanced global structural rigidity and dual-directional thermal resilience to PxSODC through a more compact hydrophobic core and a more resilient intramolecular hydrogen bond network, providing a direct structural basis for its increased catalytic efficiency at lower expression levels.

The above analyses revealed naturally occurring non-synonymous mutations in PxSODC that are enriched in the hot and cold strains. To directly test whether PxSODC is functionally required for thermal adaptation, we generated loss-of-function mutants by disrupting PxSODC in the ancestral strain using CRISPR/Cas9-mediated mutagenesis. Of 162 eggs treated with CRISPR/Cas9, 75 successfully developed into adults. We confirmed three mutant strains in the G0 generation of P. xylostella: +1 bp (SODC-MU1), +2 bp (SODC-MU2) and -1 bp (SODC-MU3). Self-crossing continued, and three types of homozygous mutation were obtained in the G5 generation (Figure 5–figure supplement 4). Life table analysis showed that the three SODC-MU strains had prolonged development, lower survival rates, and reduced fecundity and population fitness compared to the ancestral strain, particularly under hot/cold environments (Appendix 2; Figure 5–figure supplement 5; Supplementary File 3-5).

We then assessed the stage-specific survival of the ancestral strain and all three SODC-MU mutant strains (SODC-MU1, SODC-MU2 and SODC-MU3) under extreme heat stress to determine whether PxSODC loss consistently impairs thermal tolerance. At 42°C, the survival rates of eggs, 3rd-instar larvae, female adults and male adults of the mutant strains were significantly lower from those of the ancestral strain at several time points (Figure 5C). However, survival rates of the mutant pupae exposed to the high temperature (43°C) were not significantly different from those of the ancestral strain at different time points (Figure 5C). The pupal stage appeared more tolerant to high temperature than other life stages, as the PxSODC knockout did not significantly reduce pupal survival at 43°C while it significantly reduced survival of eggs, larvae, and adults at 42°C (Figure 5C). This may be due to pupal-specific heat resistance mechanisms, such as protective chrysalis and U-shaped metabolism (Kaiser et al., 2010; Barros-Cordeiro et al., 2014). In addition, supercooling and freezing points of the mutant strains (MU1: -21.32±0.41 and -9.75±0.38; MU2: -21.50±0.38 and -9.93±0.43; MU3: -21.23±0.48 and -9.94±0.41) were significantly higher than those of the ancestral strain (-23.09±0.26 and -11.58±0.52) at pupal stage (Figure 5D), indicating a key role of the PxSODC gene in the adaptability and tolerance of P. xylostella to extreme temperatures.

To further investigate the effect of PxSODC gene mutations on the temperature adaptability of P. xylostella, we identified five genes from the same SOD family in transcriptomes of the 3rd-instar larvae from the three tested strains. We found that Px04C00505 and Px13C00423 showed SNP mutations in the hot and cold strains, whereas Px20C00248, Px15C00224 and Px15C00223 were not mutated (Supplementary File 6). Further comparison of gene expressions across different strains revealed that, relative to the ancestral strain, the expression levels of PxSODC, Px04C00505, and Px13C00423 were significantly reduced in the hot and cold strains, while the remaining genes maintained stable or increased expression levels (Figure 6A). Concurrently, SOD activity decreased in the hot and cold strains, along with a reduction in O2- levels (Figure 6B-C). When SOD expression and activity, as well as O2- levels, were compared under different temperature conditions between the ancestral and hot/cold strains, similar patterns were observed (Figure 6D-F). These results suggest that non-synonymous mutations in the hot and cold strains may alter SOD protein conformation, increasing catalytic efficiency per molecule and enabling effective O - scavenging at lower expression levels. This energy-efficient strategy is beneficial under thermal stress, where conserving metabolic resources for development and reproduction is critical for survival. Compared to the ancestral strain, expression of the mutated SODC genes (Px04C00505 and Px13C00423) was increased in the male adults in SODC-MU1 and SODC-MU2. The expression levels of the non-mutated genes Px20C00248, Px15C00224 and Px15C00223 were also increased (Figure 6G). Further, SOD activity decreased, while O2-levels increased in the two mutant strains (Figure 6H-I), which were unable to fully compensate for the effects caused by the deletion of the PxSODC gene, implying that the SOD protein encoded by PxSODC plays a crucial role in O2- scavenging.

SOD expression and activity and superoxide anion (O2-) levels across developmental stages and temperature environments in different strains of P. xylostella.

(A) Expression levels of the SOD genes at different developmental stages of AS, HS, and CS in the favorable temperature environment (26°C). (B) SOD enzyme activity at different developmental stages of the ancestral, hot and cold strains (AS, HS, and CS) in the favorable temperature environment. (C) O - levels at different developmental stages of AS, HS, and CS in the favorable temperature environment. (D) Expression levels of the genes from the SOD family in the 3rd-instar larvae of AS, HS, and CS in the hot (32°C, 34°C, 36°C) and cold (12°C, 10°C, 8°C) environments. (E) SOD enzyme activity in the 3rd-instar larvae of AS, HS, and CS in the extreme temperature environments. (F) O2- levels in the 3rd-instar larvae of AS, HS, and CS in the hot and cold environments. (G) Expression levels of SOD family (excluding the PxSODC gene) at different developmental stages of the ancestral and SODC-MU strains in the favorable temperature environment. (H) SOD enzyme activity at different developmental stages of the ancestral and SODC-MU strains in the favorable temperature environment. (I) O - levels at different developmental stages of the ancestral and SODC-MU strains in the favorable temperature environment. n = 3 biologically independent samples in (A), (D), (G); within each of the boxes, the numerical value represents log2-fold change of the gene expression level in the treated samples with respect to the control. n = 4 biologically independent samples in (B), (C), (E), (F), (H), with data being presented as mean±SEM. One-way ANOVA with Tukey’s test was used for comparison in (A), (B), (C) and (G), (H), (I) (p < 0.05). t-test was used for comparison in (D), (E), (F) (p < 0.05).

PxSODC-allied metabolic networks

Untargeted metabolomic analysis of the ancestral and SODC-MU strains across developmental stages revealed broad metabolic adjustments involving lipids, nucleotides, carbohydrates, and amino acids following PxSODC deletion (Appendix 3; Figure 7-figure supplement 1 and 2). In the metabolome, the abundance of three metabolites, namely 5-hydroxymethyluracil, 2-methylcitric acid, and 5’-deoxyadenosine, differed between the mutant strains (SODC-MU1 and SODC-MU2) and the ancestral strain across all developmental stages (Figure 7A). With the exception of 5’-deoxyadenosine, which was significantly higher in the 3rd-instar larvae, these three metabolites were significantly lower in other developmental stages of the mutant strains (Figure 7B). This suggested a possible direct association with PxSODC, and may represent a key biological regulatory response in P. xylostella’s adaptation to different environmental conditions. Importantly, this specific metabolic signature provided a compelling, data-driven hypothesis linking PxSODC-mediated oxidative stress management directly to epigenetic regulation: 5-hydroxymethyluracil is directly involved in dynamic DNA demethylation, and 5’-deoxyadenosine is a precursor to S-adenosylmethionine, the primary methyl donor for DNA methylation (Bhutani et al., 2011; McKean et al., 2019). Their consistent alteration in the SODC-MU strains suggested a potential link between PxSODC and DNA methylation. To test this hypothesis, we measured the expression levels and enzymatic activities of DNA methyltransferase 1 (PxDnmt1) in the 3rd-instar larvae of different strains. The results showed that both the expression and enzymatic activity of PxDnmt1 were significantly reduced in the hot and cold strains compared to the ancestral strain (Figure 7C-D). Using RNA interference (RNAi) technology, we specifically silenced the expression of PxDnmt1 in the ancestral strain of P. xylostella (Figure 7E) and observed significantly reduced levels of 5-methylcytosine (5-mC, a marker of DNA methylation) in both pupae and female adults (Figure 7F) (Ni et al., 2023). Further, we found that silencing of PxDnmt1 could significantly decrease the critical thermal maximum (CTmax) of female adults and increase the supercooling and freezing points of pupae (Figure 7G-H). These results suggest that DNA methylation may play a role in the thermal tolerance of P. xylostella, though further work is needed to establish a direct mechanistic link between PxSODC, methylation-related metabolites, and epigenetic regulation of thermal adaptation.

Comparison of metabolites and DNA methylation across different strains of P. xylostella.

(A) A Venn diagram showing the metabolites that are consistently different between the ancestral and mutant strains across different developmental stages. (B) Three metabolites with persistent discrepancy across different developmental stages in the ancestral and mutant and strains. (C) Expression level of the DNA methyltransferase 1 gene (PxDnmt1) in the ancestral, hot and cold strains. n = 17 biologically independent samples. (D) DNA methyltransferase activity in the ancestral, hot and cold strains. n = 12 biologically independent samples. Data are presented as mean±SEM, one-way ANOVA with Tukey’s test was used for comparison in (C), (D) (p < 0.05). (E) Injection of dsDnmt1 significantly reduced the expression level of PxDnmt1 in the ancestral strain of P. xylostella. n = 3 biologically independent samples. (F) Silencing of PxDnmt1 decreased 5-methylcytosine (5-mC) content in the female adults and pupae of P. xylostella. n = 4 biologically independent samples. (G) Female adults with silenced PxDnmt1 exhibited a significantly decreased critical thermal maximum (CTMax). n = 20 biologically independent samples. (H) Pupae with silenced PxDnmt1 displayed elevated supercooling and freezing points. n = 40 biologically independent samples. Data are presented as mean±SEM, unpaired t-test was used for comparison in (E), (F), (G), (H) (p < 0.05).

Discussion

Insects are valuable bio-indicators of the effects of climate change via their phenology, distribution, population dynamics responses (Bale et al., 2002; Halsch et al., 2021). The present study demonstrates that P. xylostella can undergo rapid genetic adaptation to thermal extremes, providing mechanistic insights into how this globally invasive pest may expand its range under climate change through coordinated transcriptomic, metabolomic, and life history changes. By conducting the life history trait, demography, and fitness assessment of the ancestral, hot and cold strains of P. xylostella, we observed that the hot and cold strains had evolved significant genetic differences from the ancestral strain in multiple traits related to thermal tolerance and fitness. In the context of global warming, P. xylostella may evolve greater flexibility across its range and ecological niche leading the hot-evolved populations to be able to persist in regions with increased temperatures due to climate change (Chen et al., 2021). With climate change, cold-adapted P. xylostella may be favored during episodes of late frost in the spring or early frost in the autumn in temperate regions. The lower supercooling and freezing points of the cold strain pupae facilitate the survival of insects in cold climates, extending their ecological adaptability to low-temperature environments (Block, 1997). The demonstrated capacity of P. xylostella to adapt to extreme thermal conditions of both forms implies that P. xylostella may survive under a broader range of climatic conditions, posing new challenges for the management and control of this worldwide pest.

Our findings reveal a significant metabolic adjustment in P. xylostella during its adaptive evolution to both high and low temperatures. The convergent reduction of lipid-related metabolites such as octadecenoic acid, epoxystearic acid, and carnitine in both the hot and cold strains (Figure 2F) suggests a shared metabolic adjustment during thermal adaptation. Lipids play a key role in energy storage and membrane stability in insects, and downregulation of lipid metabolism may enable P. xylostella to conserve metabolic resources and reallocate energy toward development and reproduction (Rommelaere et al., 2019; Mallard et al., 2018). As a key mediator of fatty acid transport and energy metabolism, the reduction in carnitine levels may further reflect this energy reallocation strategy (Bremer, 1983).

Although gene expression between the hot and cold strains does not show a strong correlation, KEGG analysis indicates that differentially expressed genes in both adapted strains share similarly enriched core pathways (Figure 3E). Notably, the convergent reduction in lipid metabolism represents a shared energy reallocation strategy. Freeing up resources to fuel the accelerated development, higher fecundity, and extreme-heat survival in the hot strain, while facilitating the extended male longevity and lower supercooling and freezing points required in the cold strain (Tigano et al., 2020; Sherpa et al., 2022; Li et al., 2024a). Relative to the ancestral strain, the cold strain exhibited more differentially expressed genes (2029 vs. 1364) but fewer differential metabolites (37 vs. 77) than the hot strain. This pattern contrasts with findings in Drosophila, where thermal adaptation produced a simpler transcriptomic (Mallard et al., 2018). This divergence in omics profiles strictly aligns with their distinct physiological requirements. The cold-adapted strain relies on broader transcriptional reprogramming to structurally maintain basic cellular homeostasis and sustain cold hardiness over its prolonged lifespan, whereas the hot-adapted strain utilizes targeted gene expression changes combined with broader metabolic rewiring to actively sustain rapid energy turnover and its accelerated life cycle.

Mutations in these genes during adaptive evolution in response of thermal adaptation can lead to novel phenotypic traits, such as changes in the life history and population fitness of insects (Gibson et al., 2019). The PxSODC gene encodes a superoxide dismutase (SOD) that scavenges superoxide anions in cells, maintaining redox balance and protecting cellular structures (Sheng et al., 2014). Under extreme temperatures, insects can adjust their survival strategy, allocating more energy to maintain fundamental life functions (Hoffmann and Sgro, 2011). Here, the deletion of the PxSODC gene led to reduced SOD enzyme activity and increased O2- levels, reducing the tolerance of P. xylostella to extreme temperatures. These findings provide novel insights into how genetic variation translates into phenotypic variation (You et al., 2024), and the ways in which P. xylostella responds – and has responded – to a changing climate (Chen et al., 2021).

Our molecular dynamics simulations provide direct structural evidence that the non-synonymous mutations enhance PxSODC protein thermostability (Figure 5–figure supplement 3). Specifically, the mutant protein maintained structural integrity under both heat and cold stress through a more compact hydrophobic core and a more resilient hydrogen bond network. This enhanced dual-directional thermostability likely increases catalytic efficiency per molecule, enabling effective superoxide scavenging at the lower expression levels observed in the evolved strains and reducing the energetic cost of enzyme production, consistent with previous findings that specific structural mutations can directly increase SOD enzyme activity (Sheng et al., 2013). After long-term adaptation, insects may acquire the ability to maintain cellular homeostasis in new thermal environments by reducing basal metabolism and allocating more energy to development and reproduction (Mallard et al., 2018). At 34°C and 36°C, the trends in SOD enzyme activity and PxSODC gene expression differ between different temperature-evolved strains, suggesting the involvement of additional genes in the regulation of SOD enzyme activity. This hypothesis is supported by our transcriptomic analysis which identified additional SOD genes (Supplementary File 5).

In the SODC gene family, two SOD genes (Px04C00505 and Px13C00423) underwent non-synonymous mutations and showed reduced expression in the hot and cold strains at different developmental stages, while the expression of three SOD genes without mutations remained relatively stable or increased at most developmental stages. The evolution of protein functions is driven by mutations, which in some cases can switch directly from one function to another through single amino acid changes (Nobeli et al., 2009; Patsch et al., 2024). Under extreme temperatures, the expression of these genes in the 3rd-instar P. xylostella larvae of the hot and cold strains trends to be similar to those of the ancestral strain. This indicates that adverse environmental conditions increase intracellular oxidative stress, which requires regulation of SOD expression and enzyme activity to scavenge superoxide anions (Islam et al., 2022). Maintaining high levels of SOD enzyme activity requires additional energy, placing a strain on cellular energy metabolism and resource allocation (Emre et al., 2013). SOD genes with non-synonymous mutations, like PxSODC, can lead to the change in protein structure or function, affecting enzyme activity and allowing for faster O2- scavenging at lower transcript levels, reducing resource requirements. The three unmutated SOD genes, if mutated, might adversely affect the moth or have other functions, such as involvement in cellular signaling pathways (Mondola et al., 2016).

Metabolomic analysis of different developmental stages in the ancestral and mutant strains revealed that after the loss of PxSODC gene, the metabolism of P. xylostella underwent temperature-adaptive adjustments involving lipids, nucleotides, carbohydrates, coenzymes and vitamins, and amino acids. This study also showed that DNA methylation plays a key role in the temperature adaptation of P. xylostella. While DNA methylation may be associated with gene activation, its main function remains the inhibition of gene expression (Stroud et al., 2015; Wang et al., 2018). Transcriptomic analysis also showed that more genes were up-regulated in the hot and cold strains compared to the ancestral strain, highlighting the role of DNA methylation in regulating gene expression to help P. xylostella maintain physiological functions and survive. In addition to directly regulating the expression of temperature responsive genes, epigenetic effects can also indirectly affect the response of insects to temperature challenges (Reynolds, 2017). By adding the molecular data of epigenetic markers, underlying mechanisms on the adaptive adaptation can be more easily elucidated. Therefore, further work is required to better understand the impact of non-genetic effects on adaptation to future climates including how they interact with genetic adaptive capacity. While our experimental evolution approach successfully uncovered potential genetic mechanisms of thermal adaptation, its ecological relevance warrants consideration. Exposing the insects to extreme 12 h thermal bounds was necessary to impose strong directional selection and observe adaptive evolution over a relatively short timeframe (∼3 years) in the laboratory. However, we acknowledge that comparing fluctuating stressful environments to a constant favorable environment (26°C) may conflate adaptation to absolute temperature extremes with adaptation to thermal fluctuation itself. Future field-based studies across natural gradients, or using fluctuating optimal control regimes, will be valuable to validate the ecological impact of these mutational and epigenetic pathways in wild habitats.

This study elucidates the molecular mechanisms underlying adaptation of Plutella xylostella to both high- and low-temperature environments and functionally validates differentially expressed genes identified in ancestral, hot-evolved, and cold-evolved strains. Nevertheless, thermal adaptation in arthropods may engage distinct, temperature-specific biological pathways; accordingly, future work will prioritize the characterization of strain-specific differentially expressed genes. Beyond functional validation of the canonical stress-associated gene PxSODC, additional genes harboring nonsynonymous mutations warrant detailed investigation to clarify their roles within the broader regulatory network. Importantly, our findings also underscore a critical role for DNA methylation in thermal adaptation in P. xylostella. More broadly, future research should integrate genetic, epigenetic, and metabolic approaches across diverse taxa to determine whether adaptive mechanisms such as SOD-mediated oxidative stress regulation and DNA methylation represent general strategies by which arthropod pests adapt to novel thermal environments and expand their ranges under climate change.

Materials and methods

Insects

The founding population of P. xylostella was established from field-collected larvae on cabbage (Brassica oleracea var. capitata) in organic farms in Fuzhou, Fujian Province, China (26°05’N, 119°18’E) in July 2012. Fuzhou experiences a subtropical monsoon climate, where summer high temperatures exceed 32°C and winter lows can drop below 10°C, making our experimental selection temperatures ecologically relevant extremes for this population. The collection site was confirmed through farmer interviews and local agricultural records to have no history of insecticide application for at least five years. Approximately 300-500 larvae were collected from multiple host plants distributed across a 2-hectare area to maximize genetic diversity and minimize founder effects. The field-collected population was reared in the laboratory under controlled conditions of 26°C and 60% relative humidity with a 12-h light/12-h dark cycle, without exposure to insecticides. This setup was referred to as the favorable temperature environment based on a previous study on the relationship between temperature and the P. xylostella development (Liu et al., 2002). Population size was maintained at >500 individuals per generation to minimize inbreeding and genetic drift. This laboratory-acclimated population was designated as the ancestral strain and served as the baseline for all subsequent experiments. Eggs and larvae were reared in sterile plastic disposable Petri dishes (90 mm) with an artificial diet containing 68 g yeast powder, 20.4 g agar, 127.5 g raw wheat germ, 3.4 g potassium sorbate, 3.4 g methyl paraben, 34 g sucrose, 10.2 g powder of radish seed, 1.7 g vitamin premix, 3.4 g ascorbic acid, 3.4 mL cola oil, 0.34 mL linoleic acid and 850 mL water The adults were allowed to mate and lay eggs in disposable paper cups, and were fed with a 10% honey solution.

The previous study on the relationship between temperature and developmental rate shows that P. xylostella can survive and develop at the temperatures 32°C at the maximum, 26°C as the optimum, and 10°C for the minimum (Liu et al., 2002). To generate thermally adapted populations, we established 18 replicate populations from the ancestral strain and randomly assigned them to three thermal regimes: (1) a hot-evolved treatment with cycling temperatures of 32°C/27°C (12 h light/12 h dark), (2) a cold-evolved treatment with cycling temperatures of 15°C/10°C (12 h light/12 h dark), and (3) a control treatment maintained at constant 26°C. The hot and cold regimes used cycling temperatures to simulate diurnal fluctuations experienced in natural environments, while the control was kept at the constant optimal developmental temperature for P. xylostella (Liu et al., 2002). All other environmental conditions (humidity, photoperiod, diet) remained identical across treatments. The populations were maintained with non-overlapping generations and a census population size of approximately 500 individuals per replicate.

Prior to the start of the selection experiment, the ancestral population was maintained in the laboratory at 26°C for approximately ∼170 generations. The thermal selection experiment was then conducted for approximately three years, encompassing ∼75 and ∼15 generations for the hot and cold strains, respectively. To minimize maternal effects and ensure that observed differences were due to genetic adaptation rather than developmental plasticity, all populations were reared for two additional generations under common garden conditions (26°C) prior to phenotypic and molecular assays. All six replicate populations per treatment were used for downstream experiments, resulting in 18 experimental cohorts in total.

Development of age-stage-specific two-sex life tables

To study the life history and population fitness of the three different P. xylostella strains, age-stage-specific life tables were developed. A total of 90 randomly selected newly laid eggs from each strain were transferred to three 90 mm diameter Petri dishes (30 eggs per dish) and maintained at the favorable temperature (26°C) (Chi et al., 2020). The stage-specific number of individuals was recorded daily (food replaced every two days). During the pupal stage, each pupa was placed in a perforated 1.5 mL centrifuge tube. After eclosion, adults were transferred to 50 mL plastic cups for mating and oviposition and fed with a 10% honey solution. We continuously monitored and recorded daily oviposition and the number of surviving adults within the population until all individuals died.

For the individual life tables, 120 randomly selected newly laid eggs from each strain (i.e., 120 eggs/strain/temperature) were placed under different temperature conditions (favorable, hot, and cold environments). The eggs were individually transferred to 30 mm diameter petri dishes (artificial diet replaced every two days). The number of surviving larvae and pupae was recorded daily, and each pupa was placed individually in a perforated 1.5 ml centrifuge tube. One newly emerged adult female and one newly emerged adult male were placed in a 25 mL plastic cup and fed with a 10% honey solution for mating and oviposition. The number of eggs laid by each adult female was recorded daily until death. The longevity of both males and females was recorded.

Life history and population fitness parameters, including egg duration, larval duration, pupal duration, preadult duration, female and male longevity, oviposition days, fecundity, intrinsic rate of increase (r), finite rate of increase (λ), net reproductive rate (R0), and mean generation time (T), were calculated based on the recorded data of age-stage-specific two-sex life tables of different strains. The numerical computation was done using the TWOSEX-MSChart software (https://www.faas.cn/cms/sitemanage/index.shtml?siteId=810640925913080000) and the bootstrap method with 100,000 replications to obtain standard errors of the fitness parameters. The paired bootstrap (BT) method with 100,000 replications was used to assess pairwise differences among all strains, including comparisons among HS, AS, and CS, as well as among the ancestral and SODC-MU mutant strains under each thermal regime. A P value of less than 0.05 indicates a statistically significant difference (Chi, 1988).

Metabolomic and transcriptomic profiling

All samples for omics profiling were collected after two generations of rearing under common conditions (26°C in climate-controlled chambers) to capture constitutive, genetically based differences between the strains, avoiding confounding effects of immediate physiological plasticity that would occur under thermal stress. The 3rd-instar larval stage was selected as the most actively feeding and rapidly growing stage, where metabolic demands and energetic trade-offs critical for adaptation are most pronounced. Samples collected for metabolomic profiling included: (1) the 3rd-instar larvae from different strains (AS, HS and CS) maintained at the favorable temperature (26°C); and (2) the 2-day-old eggs, 1-day-old 3rd-instar larvae, 2-day-old pupae and newly emerged adults from the ancestral and SODC-MU (MU1 and MU2) strains (see section Deletion of the targeted gene using the CRISPR/Cas9 genome editing) at the favorable temperature. Each biological replicate consisted of pooled individuals at the same developmental stage. Six independent biological replicates per strain were used for metabolomic profiling and three for transcriptomic profiling. The same sample collection strategy was applied to both analyses, but replicates were collected independently.

Metabolomic profiling

To identify metabolic changes associated with thermal adaptation, we performed targeted metabolomic profiling using UPLC-MS/MS. Stored samples were thawed on ice and weighed (50 ± 2 mg) into 1.5 mL centrifuge tubes, to which three pre-cooled steel balls (3 mm) and 500 µL of pre-cooled 70% methanol (Merck, Germany) were added. Each of the samples was homogenized in a pre-cooled tissue homogenizer (25 HZ, 5 min) (Tissuelyser, Qiagen), and then the homogenized sample was left on ice for 15 min, then centrifuged at 12,000 rpm at 4°C for 10 min. The supernatant was transferred to a new 1.5 mL centrifuge tube and stored overnight at -20°C. The following day, the samples were centrifuged at 12,000 rpm for 3 minutes at 4°C. The supernatant was then collected with a sterile syringe and filtered through a 0.22 µm filter (Waters, USA) into an HPLC sample vial. The instrumental system for data acquisitions mainly used ultra-high performance liquid chromatography and tandem mass spectrometry (multiple reaction monitoring mode). The chromatography and tandem mass spectrometry conditions were as described by Li et al (2021). For all metabolomic comparisons (HS vs. AS, CS vs. AS, and SODC-MU vs. AS), differential metabolites were identified through pairwise comparisons using Student’s t-test with false discovery rate (FDR) correction. A multi-criteria threshold of |log2Fold Change| ≥ 1, VIP (variable importance in projection) ≥ 1, and FDR < 0.05 was applied. All differential metabolites were assigned to different pathways by KEGG analysis.

RNA extraction and cDNA synthesis

To obtain templates for gene cloning and qRT-PCR analysis, total RNA was extracted using the Eastep® Super Total RNA Extraction Kit (Promega, Shanghai) according to the manufacturer’s instructions. RNA integrity and quality were assessed using a NanoDrop 2000 spectrophotometer (GE Healthcare, USA) and 2% agarose gel electrophoresis. Total RNA (2000 ng) was reverse transcribed into cDNA using the Reverse-Transcription System Kit (Promega, Shanghai).

Transcriptomic profiling

To identify gene expression changes associated with thermal adaptation, mRNA libraries were constructed for each of the samples and sequenced on the Illumina HiSeq4000 platform (Illumina, San Diego). Raw reads obtained from sequencing were filtered, low quality reads were removed using adapters, and clean reads were obtained for subsequent information analysis. Clean reads were aligned to the P. xylostella genome using HISAT2 (V2.1.0) (http://121.37.197.72:5010), with sequence alignment performed using the software’s default parameters. Gene expression levels were measured using FPKM (fragments per kilobase of transcript per million fragments mapped). The P-value was corrected for multiple hypothesis testing following Benjamini and Hochberg (1995). Differential expression analysis between samples was performed using DESeq2. Differentially expressed genes were identified using the criteria of |log2Fold Change| ≥ 1 and FDR (false discovery rate) < 0.05. All differentially expressed genes were classified into different pathways by KEGG analysis.

Weighted gene co-expression network analysis (WGCNA)

To identify gene modules correlated with differential metabolites between thermally adapted and ancestral strains, we performed weighted gene co-expression network analysis (WGCNA). Genes with an FPKM value below 0.1 were filtered out. The WGCNA package in R (v2.0) was used to calculate weight values and a soft threshold was determined based on the scale-free network principle. A gene clustering tree was constructed based on gene expression correlations, and gene modules were identified based on these clustering relationships. The minimum number of genes per module was set to 30, and the cut height threshold was set to 0.25 to merge potentially similar modules, with other parameters set to default values. Differential metabolites shared between the hot and cold strains relative to the ancestral strain were used as trait data for correlation analysis, as metabolites represent intermediate molecular phenotypes that bridge gene expression and organismal-level traits, enabling more direct identification of functionally relevant gene modules. Pearson correlations were calculated between each module eigengene and each of the 30 common differential metabolites (29 modules × 30 metabolites = 870 tests). Following standard WGCNA practice, a stringent dual threshold of |correlation coefficient| > 0.8 and P < 0.05 was applied to identify significant module-metabolite associations, effectively controlling for false positives (Langfelder and Horvath, 2008). Genes within specific modules were compared with differentially expressed genes and those common to both were considered as candidate genes.

Gene cloning

To identify non-synonymous mutations in candidate genes across the different strains, the following gene sequences were amplified: (1) differentially expressed genes identified from specific modules; and (2) genes from the PxSODC gene family. Reference sequences for these genes were obtained from the P. xylostella genome database and full-length primers were designed using Primer Premier 6.0 (Supplementary File 7). PCR amplifications were performed using cDNA from the ancestral, hot and cold strains of P. xylostella (at least six samples) as templates, using Phanta® Max Super-Fidelity DNA Polymerase (Vazyme, China). The 50 µL PCR reaction mixture contained 2× reaction buffer (25.0 µL), dNTP mix (1.0 µL), 20 µM upstream primer F (2.0 µL), 20 µM downstream primer R (2.0 µL), DNA polymerase (1.0 µL), nuclease-free water (15.0 µL) and cDNA (4.0 µL). Amplification conditions were as follows 95°C for 3 minutes, followed by 35 cycles of 95°C for 15 seconds, 58°C for 15 seconds and 72°C for a gene-specific extension time, with a final extension at 72°C for 5 minutes. The integrity of the PCR products was verified by 2% agarose gel electrophoresis and purified using the Omega gel extraction kit (USA). The purified fragments were cloned into the pESI-Blunt vector using the Hieff CloneTM Zero TOPO-Blunt Simple Cloning Kit (YEASEN, Shanghai, China) and transformed into competent DH5α cells (YEASEN, Shanghai). The ligated product was transferred to DH5α competent cells and plated on LB+100 μg/ml ampicillin plates and incubated overnight at 37°C. A single colony was picked and placed in 500 µL of liquid LB with 100 µg/mL ampicillin. A positive clone was sent to Sangon Biotech (Shanghai, China) for sequencing. Sequence alignments of candidate genes from different strains were performed using Snap Gene software.

Sequence analysis and phylogenetic tree construction

To determine the exons and introns of the target genes, their sequences were aligned with gDNA from the P. xylostella genome database. Protein sequences were analyzed using a protein sequence analysis and classification tool (InterPro, http://www.ebi.ac.uk/interpro/). The relative molecular masses and isoelectric points of the proteins were predicted using Expasy (http://expasy.org/tools/dna.html). The secondary structures of the proteins were predicted using PSIPRED (http://bioinf.cs.ucl.ac.uk/psipred). A phylogenetic tree was inferred using the model-based Maximum Likelihood method implemented in IQ-TREE, and the robustness of the tree was verified by bootstrap analysis (bootstrap = 1000 replicates). In the absence of a valid outgroup sequence, the resulting gene tree was presented as unrooted

Molecular dynamics (MD) simulations

To investigate the structural consequences of the identified non-synonymous mutations in PxSODC under thermal stress, molecular dynamics (MD) simulations were performed using GROMACS 2022.3. The 3D structures of the wild-type (WT) and mutant (MU) PxSODC proteins were predicted using AlphaFold2. The Amber99sb-ildn force field was applied to the protein systems. Each system was solvated in a cubic box with TIP3P water molecules and neutralized by adding appropriate NaC ions. Energy minimization was performed using the steepest descent algorithm. Subsequently, 100 ps of NVT (constant volume and temperature) equilibration followed by 100 ps of NPT (constant pressure and temperature) equilibration were conducted with a coupling constant of 0.1 ps. Production MD simulations were then run for 100 ns (5,000,000 steps with a 2 fs timestep) at three temperatures: 15°C (288.15 K, cold stress), 26°C (299.15 K, favorable baseline), and 32°C (305.15 K, heat stress). Structural stability and dynamic properties were evaluated using built-in GROMACS analysis tools, including root mean square deviation (RMSD), solvent accessible surface area (SASA), and intramolecular hydrogen bond number. The 26°C simulation served as the physiological baseline for interpreting stress-induced structural changes.

qRT-PCR analysis

To validate transcriptomic data and quantify expression levels of candidate genes across strains and temperature conditions, we collected samples as follows: (1) we randomly selected 22 genes from the transcriptomes of the 3rd-instar larvae of ancestral, hot, and cold strains to validate the transcriptome data; (2) we collected samples from the eggs, 3rd-instar larvae, pupae and adult males and females of the ancestral, hot, cold and SODC-MU (MU1 and MU2) strains to assess the transcription levels of genes including PxSODC, Px04C00505, Px13C00423, Px20C00248, Px15C00224 and Px15C00223; (3) we collected the 3rd-instar larvae of the ancestral and hot strains exposed to different high temperature treatments (32°C, 34°C, 36°C) to analyze the transcription levels of the above-mentioned genes; and (4) we collected the 3rd-instar larvae of the ancestral and cold strains exposed to different low temperature treatments (12°C, 10°C, 8°C) for similar assessments.

qRT-PCR primers were designed using Primer Premier 6.0, with PxRPL32as the reference gene (Supplementary File 7). The reaction mixture of 20 μL contained 10 μL 2× Taq Pro Universal qPCR Master Mix (Vazyme, Nanjing, China), 0.4 μL of each primer, 7.15 μL nuclease-free water and 2.0 μL cDNA. The QuantStudio Real-Time PCR System (Thermo, USA) protocol was as follows: 95°C for 30 s; 40 cycles of 95°C for 10 s, 60°C for 30 s; followed by melting curve analysis of 95°C for 15 s, 60°C for 1 min, 95°C for 15 s. Each sample contained 3 biological replicates and 3 technical replicates, and the relative expression of genes was calculated using the 2-ΔΔCt method.

Deletion of the targeted gene using the CRISPR/Cas9 genome editing

To functionally validate the role of PxSODC in thermal adaptation, we generated stable homozygous mutant strains of P. xylostella with the PxSODC gene deleted using the CRISPR/Cas9 system. The target site was designed based on the 5’-N20NGG-3’ motif (underscore indicates PAM sequence), and the potential off-target effect of sgRNAs was predicted using Cas-OFFinder (http://www.rgenome.net/cas-offinder). The in vitro transcription template for sgRNA was generated from a single nucleotide strand under the following conditions: 95°C for 3 min, followed by 35 cycles of 95°C for 15 s, 68°C for 15 s and 72°C for 30 s, with a final extension at 72°C for 5 min. The amplified product was purified by gel extraction. The sgRNA was obtained by in vitro transcription of the gel-purified product using the HiScribe™ T7 Quick High Yield RNA Synthesis Kit (New England Biolabs, USA). The reaction mixture contained 2.5 μL NTP buffer mix, 0.5 μL T7 RNA polymerase mix, 65 ng gel-purified product, made up to 5 μL with nuclease-free water. After overnight incubation at 37°C, 0.5 μL of DNase was added to remove DNA and the product was incubated at 37°C for 20 minutes to yield sgRNA. The sgRNA was purified by phenol-chloroform extraction and stored at -80°C.

We prepared a 10 μL reaction mixture containing 300 ng/μL sgRNA and 200 ng/μL Cas9 protein (GenCrispr, Nanjing), 1 μL 10× reaction buffer and nuclease-free water to make up to 10 μL and incubate at 37°C for 25 minutes. The mixture was injected into freshly laid eggs using the Olympus SZX16 microinjection system (Olympus, Japan) and the entire microinjection was completed within 30 minutes of eggs being laid. After injection, the eggs were placed in a Petri dish and the number of eggs hatched was recorded. Adult gDNA was extracted using the TiANamp Genomic DNA Kit (TIANGEN, China). Specific primers were designed for PCR amplification (Supplementary File 7), with conditions as follows: 95°C for 3 min, followed by 34 cycles of 95°C for 15 s, 58°C for 15 s and 72°C for 15 s, with a final extension at 72°C for 5 min. The sequence of the PCR products was checked by Sangon Biotech (Shanghai) Co., Ltd.

The injected eggs were referred to as the G0 generation. These were reared to adulthood, crossed with the ancestral (non-injected) adults and used to extract genomic DNA from G0 adults after oviposition (the resulting progeny representing the G1 generation). PCR products flanking the two sgRNA target sites were amplified as mentioned above to determine genotypes and identify heterozygotes (individuals with double peaks in the sequence chromatogram starting from the sgRNA target site). The G1 generation was self-crossed to produce the G2 generation, and all G1 adults were genotyped based on PCR amplification for individual identification. G2 progeny derived from G1 heterozygotes with the same allelic mutation were selected. The G2 generation was then self-crossed to produce the G3 generation, retaining those with the same type of homozygous mutations to establish homozygous lines. If the G3 generation remained heterozygous, self-crossing continued until homozygous mutations were obtained (Wang et al., 2020). By the end, three mutants were obtained and called SODC-MU (MU1, MU2 and MU3) strains.

RNA interference mediated silencing of target genes

To assess the role of PxDnmt1 in thermal tolerance, we silenced its expression using RNA interference. Gene-specific primers containing T7 promoter sequences were designed (Supplementary File 7), and PCR was performed using total P. xylostella cDNA as a template. The PCR products were purified using a gel extraction kit. Double-stranded RNA (dsRNA) was synthesized by in vitro transcription using the T7 High Yield RNA Transcription Kit (Vazyme, Nanjing, China). The dsRNA was diluted to 2 μg/μL using DEPC-treated water (Beyotime, Shanghai, China). A volume of 500 nL diluted dsRNA (dsEGFP or dsDnmt1) was injected into pupae using a Nanoliter 2000 microinjector (World Precision Instruments LLC, USA). Total RNA was extracted 24 hours after injection and reverse transcribed to cDNA. Gene knockdown efficiency was analyzed by qPCR using pupae injected with dsEGFP as controls. The experiment was performed in three independent biological replicates (Zhou et al., 2024).

Assessing the response to high temperature

To assess the response of different P. xylostella strains (ancestral strain, hot strain and mutants) to extremely high temperatures, 2-day-old eggs, 1-day-old 3rd-instar larvae and 2-day-old pupae were individually placed in 90 mm diameter Petri dishes. Adult females and males were placed individually in perforated 1.5 mL centrifuge tubes. Based on a preliminary trial on the stage-specific temperature tolerance limit of P. xylostella (eggs, larvae, pupae, and both male and female adults of the ancestral and hot strains were placed in different temperature environments ranging from 40 to 45°C), pupae from the ancestral and mutant strains were exposed to 43°C while eggs, larvae, and adults were exposed to 42°C for periods ranging from 30 to 180 minutes. After treatment, all replicate samples were transferred to an environment maintained at 26°C, where survival was observed and recorded. Survival was defined as the successful development of eggs, larvae and pupae to the next stage, while adults had to show movement of an appendage or mouthparts. Experiments were performed with six biological replicates, with each replicate contained 20 individuals.

We randomly selected 20 female adults injected with dsRNA to determine their critical thermal maximum (CTMax). A thermistor probe (Omega, USA) was inserted into a 1.5 mL centrifuge tube, which was suspended inside a 50 mL centrifuge tube with the opening sealed with cotton. This assembly was then placed in a 2 L glass beaker containing 1000 mL water, with the beaker top sealed with insulating foam board. The entire setup was positioned on a thermostatically controlled magnetic stirrer, where the temperature inside the 1.5 mL tube was increased at a constant rate of 0.5°C/min. When the temperature reached 26°C, female adults were quickly transferred into the 1.5 mL centrifuge tube containing the temperature probe, and their behavioral responses were continuously monitored as temperature increased. The CTMax was recorded when moths exhibited spasms, lost their crawling or flying ability, and remained motionless at the bottom of the tube, typically lying ventral side up (in most cases) or dorsal side up (in fewer instances). Although antennae and limbs might still exhibit slight tremors at this point, the insects typically died within seconds (Li et al., 2024b).

Measurement of the supercooling and freezing points

To investigate the cold hardiness of different P. xylostella strains (including the ancestral, cold and mutant trains), we randomly selected 40 pupae from each strain to examine their supercooling and freezing points. A thermistor probe from a subcooling point tester (Omega, USA) was attached to a pupa, secured with conductive tape and placed in a centrifuge tube, with the tube mouth sealed by cotton. The centrifuge tubes were then placed in a 50 mL plastic cup filled with ethylene glycol (antifreeze), and the cup was stored in an ultra-low temperature freezer set at -70°C, with the temperature first dropping rapidly and then decreasing at a rate of 0.10°C per second until the supercooling point was reached. By recording temperature changes at intervals of every second, the supercooling and freezing points of pupae were determined based on the inflection point of body temperature. The same experimental approach was also applied to P. xylostella injected with dsRNA.

Detection of oxidative stress indicator

To assess the impact of thermal adaptation and PxSODC deletion on oxidative stress, we measured superoxide dismutase (SOD) activity and superoxide anion (O -) levels. Samples were collected in the following conditions: (1) the eggs, 3rd-instar larvae, pupae, and adult males and females of the ancestral, hot, cold, and SODC-MU (MU1 and MU2) strains at the favorable temperature (26°C); (2) the 3rd-instar larvae of the ancestral, hot, and SODC-MU (MU1 and MU2) strains after 2 hours of heat stress at 32°C, 34°C, and 36°C; and (3) the 3rd-instar larvae of the ancestral, cold, and SODC-MU (MU1 and MU2) strains after 2 hours of cold stress at 12°C, 10°C, and 8°C. The experiment was performed with four independent biological replicates. The levels of SOD and O2- were measured using commercial assay kits (Comin, China) according to the manufacturer’s instructions.

Detection of 5-methylcytosine concentration

To evaluate the effect of PxDnmt1 silencing on DNA methylation levels, pupae and female adults were collected for detection of 5-methylcytosine (5-mC) concentration after injection with dsRNA. The levels of 5-mC were measured using a commercial insect 5-methylcytosine (5-mC) ELISA detection kit (Shanghai Enzyme-linked Biotechnology Co., Ltd., China) according to the manufacturer’s instructions. The experiment was performed with four independent biological replicates.

Data analysis

Statistical analyses for life table parameters, metabolomic data, and transcriptomic data are described in their respective sections above. For all other experimental data (qRT-PCR, SOD activity, O2- levels, DNA methyltransferase activity, 5-mC concentration, stage-specific survival rates, and supercooling/freezing points), analyses were performed using SPSS version 23.0. The Shapiro-Wilk test was used to assess normality of data distribution. For normally distributed data, two-group comparisons were analyzed using independent samples t-tests, while comparisons involving three or more groups were analyzed using one-way ANOVA followed by Tukey’s test (homogeneous variances) or Tamhane’s T2 test (unequal variances). For non-normally distributed data, the Mann-Whitney test (two groups) or Kruskal-Wallis test (three or more groups) was used. A P value of less than 0.05 was considered statistically significant in all cases (Lei et al., 2024).